---
title: (working title) Control mechanisms for working with consultants
biblio-style: apalike
csl: apa.csl
blinded: 0
link-citations: yes
authors: 
- name: Roel Peters
  affiliation: Antwerp Management School
bibliography: references.bib
output: rticles::asa_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

# Defining "Digital Consultancy"

"Digital consultancy" is defined as consultancy in (either or both) technological and organizational aspects of digital transformation. The following two sections substantiate this definition by elaborating on the two concepts that comprise this definition. First, six properties that define "consultancy" are discussed, followed by an elaboration on "digital transformation."

## Consultancy

Consultants, or management consultants, have been described through a multitude of metaphors and nicknames: "capitalism's commissars" [@thrift2005 93], "shadowy figures operating in the background but exercising considerable influence" [@kipping2012 31], "agents of a modern rationalistic and universalistic culture" [@kipping2012 190], "institutionally approved agents" [@kipping2012 193], "marketized experts" [@furusten2012 265], "magical figures, shamans or witch doctors" [@fincham2002 68], "puppet masters" [@fincham2002 69], "knowledge entrepreneurs that promote emotionally charged, enthusiastic, and unreasoned discourse" [@leicht2006 37], "supra-experts" [@kieser2006 94], "improvising experts" [@furusten2009 272], "brochuremanship" [@levine1982 75] and simply "The Big Con" [@mazzucato2023].

Their work makes use of a vague body of knowledge described as "elusive", "fuzzy", "perishable", "indeterminate", "esoteric", "fluid" and "changeable" [@muzio2011]. Consequently, for buyers it is hard to know what they need and what they get, and for sellers of consultancy it is hard to know what to offer [@furusten2012 266]. Furthermore, consultancy is marked by very low professionalization since occupational entry is unprotected, the supply of labor is unregulated and there is no formal accreditation. [@fincham2006 20] Practicing "consultancy" is the main criterion of membership with competences and "time spent in the industry" as the main differentiators.

These remarks are prima facie evidence that no consensus around the definition exists, and what they do is extremely hard to describe. @kipping2012 [24] states that "definitions of management consultancy are problematic because the permeable boundaries of the industry have resulted in significant shifts over time in the composition of the industry. This means that what comprises consulting work is dynamic, ever shifting, and contested as new firms enter the industry and techniques deemed formerly appropriate, change. Although the industry is characterized by periodic structural shifts, at its heart it is an advisory activity built on the client--consultant relationship. [...] it is perhaps this chimeral ability to avoid precise definition and to be able to constantly reinvent its core services to meet ever changing understandings of the problems that beset contemporary organizations, which partly underpins its growing economic importance."

The following definition of "consultancy" is used throughout this paper.

> *Consultancy is a service offered by an external service provider. Although the responsibilities of a consultant are highly contingent on the client organization and consultancy can take many forms and require a variety of expertise, its goal is to establish change in the procedures, organizational structure or tools of a client organization. Finally, the success of a consultancy engagement is often determined by the interactivity between a consultancy firm and the client organization.*

The sections below unpack this definition and elaborate on the six properties that it comprises.

### External

@chowdhury2021 [138] describes consultants as "external advisors to corporations, nonprofits, governments, and any other forms of organizations."

### Change

Clearly, the construct of a "consultant" cannot be described by the topic that they work on, nor their academic and professional background, accreditation or membership. Instead, we should look at their goal(s): @werr1986 [1] implies that there is always a change process between clients and consultants. This is confirmed by @kipping2000 [12] who states that "management consultancies earn money through changing current procedures in client organizations." Although this change is often described by consultants as a 'tailored solution', consultants provide a service, which is inherently intangible (compared to the 'solid' nature of products) and hard to evaluate [@fincham1999 348], especially because an evaluation should not only account for the content of the changes, but also in terms of the competence development of the client, as a result of the change process [@werr1986 17].

Although consultants' goal is to establish change in an organization, their role is often symbolic. As external consultants, associated with their "quest for knowledge" and their "quest for excellence", they are [@pellegrin2006 9-13] well-equipped for legitimizing hard decisions, signaling importance and providing meaning.

Establishing change is what sets consultancy apart from temporary staffing. "A temp is generally not supposed to change the work practices at the client organization. A consultant is often expected to do just that, or at least to provide an alternative point of view." [@furusten2000 5]

The aspect of change is also important for drawing boundaries between consultancy and outsourcing. Consultants are often hired for defining a problem and presenting a solution at the same time [@furusten2009 272], while outsourcing simply delivers the solution. In that sense, outsourcing is the practice of obtaining goods or services from an external provider, as a substitute for sourcing it internally [@lacity2012 2], for a contractually agreed monetary fee and period of time [@leimeister2010 20-21]. Or in the words of @zhu2001 [374]: "the process of transferring the responsibility for a specific business function from an employee group to a non-employee group." While early IT outsourcing initiatives were rather "total" @willcocks1995 in nature, outsourcing individual business functions is now a more common activity [@zhu2001 377], and focus has shifted from cost saving to quality, productivity, flexibility and technological diversity [@kirilov2012 185]. This implies that outsourcing, unlike consultancy, is not about changing a procedure or service, but rather ensuring continuation, or implementation, by a third-party provider[^1].

[^1]: *Prima facie*, this could be interpreted as a departure from the work by @loh1992 and @venkatraman1994, who claim that IT outsourcing is an "administrative innovation, [...] involving significant changes in the routines used by the organization to deal with its tasks of internal arrangements and external alignments." However, while there is indeed a firm-level change [@nelson1985 14], in the sense of who is responsible for a specific (part of a) procedure, the procedure or service involved does not change *an sich.*

Finally, @furusten2009 [272-273] argues that consultants not only act as agents of change, but often as agents of stability. Consultants are often perceived as "relatively stable by [...] employees, stakeholders and other external counterparts", this "builds confidence for the organization. The more stable an organization is experienced as being, the greater its ability to concentrate on its core activities."

### Contingent

Being a consultant implies taking on a variety of responsibilities throughout a certain time span. Although many consultants have structured methodologies, which are converging across the industry [@werr1986 17], "[c]onsultants operate in an intense environment that regularly entails new challenges." [@chowdhury2021 138] What entails consultancy work is dynamic, ever shifting and contested with every new firm entering the industry and new methodologies claiming the spotlight [@kipping2012 24].

This is what sets external consultancy apart from internal consultancy.

### Relational

The nature of consultancy is often relational. First of all, a consultant's work is embedded in an organization's web of interpersonal relations. "[T]he context, terms of reference, and ensueing recommendations pertaining to a consultancy engagement may represent a continuation, by other means, of ongoing processes of co-operation, struggle and conflict between organizational groups." [@bloomfield1995]

Furthermore, a consultant's deliverable is often co-created with the client. @nikolova2009 [290-297] describes the client-consultant relationship within a 'social learning model'. It starts from the belief that there is no "knowledge out there", and client and consultant need to work closely together to develop problem solutions. The role of the consultant is that of a "facilitator of diagnosis and problem-solving" and coach, while the client is the actual problem solver. @clark1998 [22] even goes so far as to claim that "Like a bottle of wine, a restaurant meal, or a book, the quality of a management consultancy service is determined during enactment/consumption. This indicates that the outcome of a consultancy service is highly dependent upon the quality of the interaction between the client and the consultant."

A consultant needs to develop skills in order to fit in and adapt their skill set to the needs of a client. When a relationship does not succeed, no authority is placed on the skills of the service provider [@furusten2000 10] and the assignment might turnout to be unsuccessful. In other words, a consultant is good at improvising: "when they do not really know what problem the client has or how to solve it, to improvise and act in a manner they believe the other party expects of them in a particular situation is a convenient way for both parties to muddle through." [@furusten2009 270]

The importance of the relational aspect again sets consultancy apart from outsourcing [@kipping2012 171-173]. The outcome of a consultancy assignment often depends heavily on the interaction between the client and the service provider. On the other hand, outsourcing focuses more on technical capabilities and implies an integral handover of a (set of) service(s) to an external provider that becomes the sole responsible for delivering them. Nevertheless, one could argue that the implementation phase of a consultancy engagement offers the same benefits as outsourcing certain IT functions. For this reason, where appropriate, this study also uses insights from literature that focuses on IT outsourcing.

Finally, the relational aspects also sets consultancy apart from auditing, for which "[i]ndependence is necessary to prevent auditors from biasing their opinions in favor of their clients." [@bazerman2011 310] In other words, to prevent client pressure leading to client pleasing by the auditor [@koch2017], audit engagements shouldn't be relational, on the contrary.

### Two-sided

Besides their tasks at the client, consultants also face internal pressures from their employers in terms of optimized resource utilization (billabillity), using proprietary knowledge and "proximity that they can develop with the client." @chowdhury2021 [138] In other words, a consultant serves two masters: the client organization, and the organization that pays their wage. While the former expects an adequate service, the latter has a commercial motive. [@furusten2012 270]

Two-sidedness is what sets external consultants apart from internal consultants, but also from temporary staff. If a worker or contractor is under the direct supervision of the organization it is working for, it is seen as temporary staffing or "staff augmentation." [@hodosi2019 1]

### Diverse

Within this group, however, we can identify consultancy types: strategy consulting, tax consulting, HR consulting, risk & regulatory consulting, etc. However, @armbruster2006 [71-72] argues that the boundaries between consultancy service types are blurred. A single project often requires multiple types of services, but the distinction is often artificial. Especially the boundary between strategy and IT consultancy is opaque due to the fact that the big accounting & strategy firms entered the IT consultancy market to conduct all-encompassing projects where strategy and IT meet.

## Digital Transformation

@bloomfield1995 [28] describes IT consultants as intermediaries: "they interpose themselves between IT and clients, or between IT suppliers and clients, in effect seeking to speak for technology. Put another way, they seek to portray them selves as obligatory passage points. [...] the problem of choosing a particular functional system is translated into a problem of choosing the best expert advice."

However, the offering by many different types of organizations of some kind of IT consultancy for selecting, implementing, configuring and preaching IT solutions has led to a blurring of consultancy work [@bloomfield1995 31; @kipping2012 162]. While consultants working for hardware and software vendors are claimed to be motivated by the sale of their own products, consultancy firms aim to equal themselves as suppliers of objective business advice.

These blurring lines are the result of the fact that there is more than a technical dimension to IT solutions. Consequently, consultants frame IT solutions not just from their technical dimension, but from their organizational dimension, as well [@bloomfield1995 24-25]. Like strategy, technology, often depicted as neutral and separate from social or political matters, can be wielded for political purposes. However, the boundary between the merely technological and political is flexible: a social or political problem can be translated as a technical one.

In accordance with this interpretation, the services of multinational consultancy firms are defined or classified as consultancy in digital operations (PwC); digital commerce & engineering (Accenture); digital transformation (EY, Bain, Deloitte, Tata); digital (McKinsey, KPMG); digital, technology & data (BCG); digitalization (Capgemini), digital solutions (BoozAllenHamilton) and digital experience (Cognizant). Ergo, it is remarkable that many research describes this group of consultants as "IT consultants" [@nevo2007; @loh1992; @fincham2006; @armbruster2006; @bloomfield1995; @schwarz2005], while none of the big consultancy firms offer "IT consultancy". @czerniawska1999 [96] abandons the term and trades it in for "IT-related consultancy".

Mindful of these findings, this paper trades in the concept of "IT consultancy" for "digital consultancy", and defines it as follows.

> *Digital consultancy is consultancy in (either or both) technological and organizational aspects of digital transformation*.

In this definition "digital transformation" refers to the expectation that the use of digital technology will lead to favorable business outcomes [@wessel2020 104-118], by redefining or supporting the value proposition of an organization and imposing changes on the work practices of its organizational members. Proceeding forward, this interpretation rolls up the dichotomy between the concepts of "digital transformation" and "IT-enabled organizational transformation" into a single term.

# The Emergence of Digital Consultancy

According to @mazzucato2023 [46] historians put the birth of modern consulting in one of the three following periods.

-   The late 19th century, with the appearance of "consultant engineers" in Europe and the United States.

-   The increasing popularity of Frederick Taylor's ideas of "scientific management" and in the early 20th century.

-   The development of cost accounting methods in the 1920s and the rise of McKinsey.

However, it was the deregulation of the 1980s that acted as a catalysator for the consultancy sector's growth. According to @mazzucato2023 [61] "neoliberalism created fresh possibilities for expansion across business and government. In the private sector, the emergence of unprecedentedly huge companies that resulted from the mergers, acquisitions and easier access to credit led management consultancies to develop advisory arms specializing in multinational strategy. Because consultancies had established offices around the world, many leaders took their claims to local expertise at face value."

@furusten2000 [3] is less pejorative: the post-bureaucratic organization simply "invites market dynamics into what used to be intra-organizational matters and seeks to rid the organization of activities that are not directly linked to its focal service or product."

Besides private companies, consultancy firms also found their way into the government. @ylonen2019 [242] use the term 'consultocracy'; the "phenomenon in which often short-term, outsourced expert knowledge production is increasingly replacing the long-term work of civil servants and even politicians. This results in an increased power of consultants over politics, public governance, and public sector practices. The calculations in @saintmartin2017, suggest that smaller public sectors consume more management consulting services than in countries where the state plays a larger role in the economy and society.

On a micro-scale, laissez-faire polices, deregulation and globalization increased "insecurity regarding the basic assumptions, discourse and practices used in describing reality." [@pollner1991 370] While markets are supposed to be the guiding signal, they "do not speak very clearly nor do they provide elite managers with clear guidance regarding the directions markets are heading in specific contexts." Seen in that perspective, consulting is the profession that "interprets markets for you". [@leicht2006 37]

In that same period, managers began to employ IT to align internal processes with their organizations' business objectives, clearly pointing to an alignment of IT and strategy. However, it were audit and accounting firms that succeeded in occupying the space. First, an audit allowed it to enter a client's business, and when they detected aspects of their systems that could be improved, there would be an opportunity for selling IT services to fix the client's problems. [@mazzucato2023 73] Next, they had accounting-related experience with large-scale data processing [@armbruster2006 121]. Third, auditing requirements had consultants validate the functioning of the software, providing them the necessary experience. Fourth, they were more used to a hands-on role while management consultancy firms "are constantly exposed over their perceived reluctance to be involved in the implementation of their recommendations." [@czerniawska1999 168] Finally, due to antitrust regulation in mature markets such as the UK and the US, big IT firms were initially barred from offering consultancy services. However, a more laissez-faire approach to antitrust by the American government, allowed large IT providers such as IBM, HP and Siemens to also expanded into consulting, as this market offered higher margins than their original hardware and software businesses. Compared to audit and accounting firms, however, they were rather late to the table.

The demand for consultants familiar with IT was amplified as new applications such as ERPs, CRMs, and accounting software hit the market. IT became recognized as a facilitator of change and IT consultants were working on client's operations and often served as change managers. "Where once a company would have spent considerable in-house time on developing their own tailor-made software, most accept that they can buy a ready-made package from a professional software house [and] accept that it is probably quicker, cheaper, and ultimately more effective to adapt [their] processes to a given package." [@czerniawska1999 23]

More recently, the introduction of network computing and the internet fundamentally transformed the nature of commercial transactions. At this point in time, IT consulting becomes digital consulting, as IT is no longer exclusively used for increasing efficiency in internal operations, but transforming "all sectors of the economy [as] computers and the internet are producing rapid changes in how goods and services are produced, the nature of the goods and services being offered, and the means by which goods and services are brought to the market." [@brynjolfsson2000 2-14]

This sparked the rise of a host of 'dot.com consultancies' that provides advice on how to exploit these new opportunities. It is in this phase that many digital consulting & outsourcing services became standardized, leading to rapid commoditization and lower prices. Think of the development of APIs, websites or mobile applications, using well-documented open-source frameworks or proprietary technologies. New companies began to offer these services on a global scale, often driven by a relatively cheap labor force in emerging economies.

Concluding this section, the following three phases can be identified: (1) IT consultancy by accounting firms between the 50s and 70s, (2) huge software projects such as ERPs fuelling the spectacular rise in demand for IT consultants, offered by both accounting forms and IT firms and (3) the internet offering opportunities for global expansion to all types and sizes of consultancy firms.

# Why digital consultancy: Practical perspectives

The question of "why" encompasses two relevant questions:

1.  What are the tasks of consultants?
2.  What are the reasons for choosing consultants over internal employees?

## Tasks of consultants

An end-to-end consultancy assignment involves many steps. The following overview by @turner1982 involves eight steps and demonstrates how consultancy is external, contingent, relational and involves change.

1.  Provide requested information.
2.  Provide a solution to a given problem.
3.  Conduct diagnosis that may redefine problem.
4.  Provide recommendations.
5.  Assist with implementation.
6.  Build consensus and commitment.
7.  Facilitate client-learning.
8.  Improve organizational effectiveness.

The fifth goal (implementation) is not without controversy as traditionally, some argued that "one who helps put recommendations into effect takes on the role of manager and thus exceeds consulting's legitimate bounds." [@turner1982] Also, "a frequent dilemma for experienced consultants is whether they should recommend what they know is right or what they know will be accepted." Finally, the author notes that the last three steps are seen as by-products, and often not as explicit goals.

If we go from consultancy, in general, to digital consultancy, it is essential to keep the scope of the definition in mind. Digital consultancy has a much broader scope than merely technological advice and implementation. @swanson2010 [20-25][^2] has described five different ways how consultants can contribute to an organization's innovation process through information technology:

[^2]: @swanson2010 utilizes the term "IT consultancy." However, by ascribing it to both IT-technical and organizational aspects of IT, it inherently refers to "digital consultancy."

-   *Business strategy*: IT consultancy can lead the organization to new pursuits and technologies they wouldn't have discovered themselves. Second, IT consultancy can frame the need for innovation in strategic terms, and they prepare and legitimize the need for change.
-   *Technology assessment*: IT consultancy can facilitate the comprehension of IT technologies and its alternatives.
-   *Business process improvement*: Innovations that involve IT usually come to fruition only after business processes have been revamped. Business process changes usually require an outside-in view and offer rich opportunities for consulting.
-   *Systems integration*: In many cases, introducing a new technology requires that it needs to be integrated with existing systems and users need to be onboarded. This type of IT consultancy usually requires coding skills, hands-on design and implementation expertise
-   *Business support services*: Finally, once the implementation is completed, it can take a while before the solution is entirely assimilated. IT consultants can provide complementary IT services such as support and maintenance until the technology is entirely embedded in the organization.

With regards to the innovation process, @bessant1995 [101-102] describes four ways that consultants can play a role: (1) direct transfer of knowledge through experts, (2) sharing experiences from other organizations, (3) a single point of contact for accessing a wide range of experts and (4) defining the particular needs of a client.

## Choosing consultants over employees

The following sections follow an amended classification by @lacity1994 and rely on research both in (IT) outsourcing and consulting, since many conclusions apply to both. In these situations, it is referred to as working with a "third party." Nevertheless, distinctions are mentioned wherever extrapolating conclusions from the former to the latter is inadequate.

## Financial Expectations

Financial expectations regarding digital consultancy and IT outsourcing are twofold: reducing costs and improving financial control.

### Cost Reduction

The expectation of cost reduction comes from the ability to save on human resources, the ability to eliminate them in times of recession, and not having to pay dues when a new technology needs to be explored and adopted. @lacity1994 [10] found that managers expect that "unit costs are less expensive because of mass production efficiencies and labor specialization." The former applies to outsourcing, but the latter apply to both. For IT outsourcing, it also involves the elimination of large fixed costs during recession and transferring adjustment cost to a third-party [@aubert1996 52].

### Cost Control

The second financial expectation is not necessary about reducing costs, but about controlling them. @sturdy1998 [233] describes how "executives wanting to exercise control over the management and investment of IT, but lacking the expertise." @lacity1994 [10] states that "[i]n most organizations [...] IT costs are controlled through general allocation systems that motivate users to excessively demand and consume resources." No surprise that involving third parties is seen "as a way to contain costs because vendors implement cost controls that more directly tie usage to costs. In addition, users can no longer call their favorite analysts to request frivolous changes but instead must submit requests through a formal cost control process."

@ketler1993 [454] found that some managers see outsourcing as a means of sharing (financial) risks, reducing potential weaknesses in a department. Problems (and associated costs) with staffing, technology and selection are transferred to, or shared with the partner. However, the author also states that other managers rather fear the risk of quality loss, which hints at some kind of trade-off.

Finally, @lacity1994 describes how some managers "wanted to use outsourcing to restructure their IT budgets from lumbering capital budgets to more flexible operating budgets."

## Business Expectations

### Developing Strategy

See @sturdy1998

### Focusing on Core Activities

@furusten2009 [272-273] argues that consultants are often perceived as "relatively stable by [...] employees, stakeholders and other external counterparts", this "builds confidence for the organization. The more stable an organization is experienced as being, the greater its ability to concentrate on its core activities."

### Facilitating Mergers & Acquisitions

Oftentimes, "[m]ergers and acquisitions create many nightmares for IT managers, who are required to absorb acquired companies into existing systems." [@lacity1994 12] Although managers expect that involving third parties for merging IT functions could solve technical incompatibilities and absorb additional employees, the authors found that this was rarely successful.

## Technical Expectations

### Hiring Skilled Individuals

@lacity1994 [12] describes that in many organizations, there is dissatisfaction with in-house IT departments delivering systems late and over budget. In that context, involving third parties is seen as a way to improve technical service. @sturdy1998 [233] agrees and states that managers rely on consultants when their departments are "lacking the skills for a project" or they want to have them "compete with each other." Nevertheless, @wood1996 [656] discovered that "consultancies tend to reinforce the strategic strengths of experienced companies rather than compensate for the weaknesses of the inexperienced."

@aubert1996 [52] found that as specialized firms have digital services as their core business, which is a source of increased efficiency and productivity. E.g. specialized firms can attract highly skilled professionals which are in short supply in the market as a whole. @mazzucato2023 [28] describes how the British National Health Service (NHS) heavily relied on data, digital and project management consultants, because these "were in short supply in the civil service."

The sought-after skills are not always technical in nature. More often, it's about being up to date, and spotting emerging trends. @werr2002 [53] found that many organizations expect consultants "to interpret the meaning of technological developments, industry changes and emerging management concepts [...] to the client organization."

@ketler1993 [452] offers another interesting perspective: as the scope of digital services expand, it is difficult or unnecessary for small firms to have a (full-time) expert in each area. Specialized third-party firms, on the other hand, offer a variety of skills and technical knowledge to their clients.

### Knowledge Transfer & Diffusion

In @werr2002 [53-54], a manager describes how they are often caught up in day-to-day activities, and consultants can help them take a loot at the "big picture", from a strategic perspective: they make sense of the manager's organization in relation to its environment, such as its competitors. The research describes it as follows: "Managers thus viewed consultants as a way of bypassing the knowledge filters created by the organizational hierarchy, as well as the effects of organizational politics, which became salient in times of reorganization and change. Management consultants were seen as a way for managers to gain a 'true' picture of what was going on in their organizations."

In thas sense, @furusten2009 [269] describes a consultant as a "carrier": "Carriers of experience, expertise, knowledge, information and data about leadership, management, organization, top-down strategies and holistic perspectives."

However, in @sturdy2009 [633], two notions of knowledge transfer are proposed. In the static notion, knowledge is hard to transfer and diffuse because of its stickiness at the source. In the action-oriented view, knowledge is made by the context (the boundaries) and constructed through interaction or practice. These boundaries can be of physical, cultural and political nature.

@cohen1990 [128-129] agrees with the latter notion: "the ability to evaluate and utilize outside knowledge is largely a function of the level of prior related knowledge [such as] basic skills, or even a shared language but may also include knowledge of the most recent scientific or technological developments in a given field. [...] These abilities collectively constitute what we call a firm's *absorptive capacity*."

@fincham2002 [84] made an interesting observation: organization-specific knowledge and expert knowledge are very complexly related, and knowledge transfer can only happen into a "well-prepared ground." @nooteboom2000 [922] describes this from a transaction cost economics (infra) perspective: "[O]ne needs to make investments that are 'specific' to the relation, [...] and a certain durability of the relation is required to set up and recoup the investment." Especially tacit knowledge (impossible to codify or document) suffers from this problem, as it can only be transferred through direct interaction and with hands-on participation by the intended recipient.

@brunson1993 [41-42] is very critical of learning and knowledge transfer through consultants: "There may even be cynics in the organization who have experienced so many reforms that they have become sceptical about the very idea of reform itself as a means of solving problems or improving performance. So reforms are facilited not by learning, but by forgetfulness, by mechanisms that cause the organization to forget previous reforms or at least those of a similar content." Three mechanisms promote forgetfulness:

-   a high turnover of personnel;

-   a high turnover of top management;

-   dependence on consultants: fresh to the organization, repeating previous mistakes and rarely involved in the implementation, let alone evaluating them.

## Political Expectations

Finally, there are also reasons that are beyond the business-economical sphere. Oftentimes, individuals want to, or need to, pursue their personal goals within an organization. They might have ideas or plans, and use "the 'objectivity' and/or status of consultants to legitimate or influence a course of action." [@sturdy1998 233]

@mcfarlan1995 [14] agree, and point to "corporate culture" and "internal irritants". When an IT department can't pull of its planned centralization strategy in a decentralized organization, outsourcing the endeavor to a third-party could provide the fulcrum to overcome this impasse, since it's not associated with a specific department. The notion of a remote, efficient, experienced outsourcing partner is very compelling in that sense.

Political expectations can also come from outside an organization. The use of consultants can instill trust in shareholders and other stakeholders. As @kieser2006 [70] noted: "Companies that are held internally and externally accountable for how they 'handle' uncertainty will contract consultants as a sign of good management. Even patients who principally distrust physicians cannot avoid consulting them."

@lacity1993 [258] also identify a number of political motives, in which managers simply hired consultants to jump on the bandwagon: to react to the efficiency imperative, justify additional resources, react to positive outsourcing media reports or enhance credibility.

@werr2002 [54] describes how consultants can be used as a leading example. They are "valued for energizing the change efforts and pushing the change projects forward. [...] In supporting the realization of change projects, consultants provided methodology as well as an 'energizing example' with their own style of working."

@ylonen2019 [250] argues that the increasing reliance on consultants in the public sector results in the loss of tacit knowledge in that sector: "The old bureaucratic virtues erode as informal trust and information lose their organizational structures and channels." The culprit is usually found in cost-cutting programs that generate pressure to eliminate permanent work hours. However, it "can also be advanced as a hidden or explicit political agenda. [...] major organizational overhauls were often motivated by the desire to destruct existing organizational structures. Constant change was desirable precisely because it helped to shatter old ways of working."

# Why digital Consultancy: Theoretical perspectives

According to @armbruster2006 [3-6], the theoretical perspectives on consultancy can be broken down into two main categories and corresponding streams of literature. The first one is the functionalist view, which sees consultants as "carriers and transmitters of management knowledge." The second perspective argues that the functionalist perspective is to narrow in scope to grasp consulting projects: client-consultants interactions are open to distortions, and understanding them requires research. This is known as the critical view.

## Transaction Cost Economics

Transaction costs economics sees economic organization as a problem of contracting, i.e. organizing economic activity. The starting point is that every transaction comes with certain costs, both ex ante and ex post.

@dahlman1979 [147-148] obtains a classification of transaction costs by going through the different phases of the transaction process.

-   Search & information costs (ex ante): "Two parties [...] search each other out, which is costly in terms of time and resources. If the search is successful [they] must inform each other of the exchange opportunity [...] and the conveying of such information will again require resources."

-   Bargaining & decision costs (ex ante): "Often [...] agreeable terms can only be determined after costly bargaining between the parties involved."

-   Policing & enforcement costs (ex post): "After the trade has been decided on, there will be the costs of policing and monitoring the other party to see that his obligations are carried out as determined by the terms of the contract, and of enforcing the agreement reached."

The last type of transaction costs arises from bounded rationality: it is impossible to estimate both the costs and risks of complete contracts, or even enacting and enforcing them. [@aubert1996 53] The result is that the contractual partners often decide to leave room for adaptation and interpretation, which, in turn, increases the risk of opportunistic behavior (infra).

What follows is that the decision whether a service should be purchased in the market is the result of a comparison of the resulting costs (including transaction costs) with the costs of producing within a "hierarchy". While hierarchies coordinate the flow of materials through sequential steps by controlling it on a higher level in the managerial hierarchy, markets coordinate them through the supply and demand forces between firms [@malone1987 485].

Typically, as they acquire new resources, hierarchies can specialize, resulting in a higher productivity. However, "[a]s a firm gets larger, there may be decreasing returns to the entrepreneur function, that is, the costs of organi[z]ing additional transactions within the firm may rise. [For example, because] the entrepreneur fails to place the factors of production in the uses where their value is greatest." [@coase1937 394-395] The result is that firms increasingly are exposed to costs of internal coordination: "[E]very time that a job which was previously done by one man or one group of men is divided into two or more parts, the problem of coordinating the work of the now separated groups or individuals begins to arise." [@robinson1931 40]

In other words, as companies aim to reduce production costs, by increasing scale, they specialize, albeit increasing coordination costs. If coordination costs would not exist, organizations would grow indefinitely. The result is that blue-collar jobs disappear as production costs are reduced, while white-collar jobs, aimed at coordination, do not. [@canback1998 31-32] Mindful of this evolution, the assumption is that there is a high demand for advice and (IT) solutions that improve coordination within and between firms. These are services in which consultants are particularly well-versed. The question to ask here is: are the transaction costs for working with external consultants lower than internal coordination costs when it comes to improvement of internal coordination and knowledge production?

@canback1998 [37-44] argues it does, and starts from the three critical dimensions of transactions, a popular research topic within transaction cost economics.

*Asset specificity* describes the degree to which physical, human or site assets have a specific usage and can not be put to use for another purpose. With highly idiosyncratic transactions, market forces fail as no vendor is willing to tailor his product or service to one client, and face downward price pressure, since the latter acts as a monopsonist [@robinson1969 218-228], and no buyer is willing to put its faith in the hand of a third party at the risk of being blackmailed. The result is a bilateral monopoly. [@williamson1985 63]

That's why, according to @williamson1979 [250-253] higher asset specificity leads either to one of two forms of "relational contracting". The first form is bilateral governance in which there are "admissible dimensions for adjustment such that flexibility is provided under terms in which both parties have confidence." The second form is unified governance (i.e. internalization or vertical integration), in which "adaptations can be made in a sequential way without the need to consult, complete, or revise inter-firm agreements. Where a single ownership entity spans both sides of the transactions, a presumption of joint profit maximization is warranted."

Initially, @williamson1985 [95-96] identifies 4 sources of asset specificity, but two other sources have been added:

1.  *site specificity*, or the degree to which the successive stages of production are in close proximity to each other;
2.  *physical asset specificity*[^3], or the degree to which the physical properties of the product are unique;
3.  *human asset specificity*, or the degree to which the skills, or configuration of skills within a team, are unique to an organization's production process;
4.  *dedicated assets*, or general investments by the seller which are made with the expectation of a considerable amount of trade with one particular buyer;
5.  *brand capital specificity* [@vita2011 335], or investment in reputation that could be harmed by a vendor delivering bad quality;
6.  *temporal specificity* [@malone1987 486], or the degree to which an asset's value is dependent on it reaching the user within a certain time limit such as shipbuilding or hotel linen delivery.

[^3]: The service sector equivalent is also known as *procedural asset specificity*, or specific routines and workflows tailored to a particular transactional relationship, which are hard to modify or redeploy without value reduction. [@zaheer1995]

The second transaction dimension is its *frequency*. A one-time transaction with high asset specificity does not require a different contracting approach, because there is no subsequent phase in which the buyer/vendor can leverage his monopsony/monopoly power and stray from the initial contract. However, when the frequency goes beyond a single transaction "idiosyncratic transactions are ones for which the relationship between buyer and supplier is quickly thereafter transformed into one of bilateral monopoly." [@williamson1985 241]

*Uncertainty*. Within the context of transaction cost economics, @shin2003 [38] states that "many empirical studies show mixed and contradictory results against what transaction cost economics predicts, especially for the concept of uncertainty." As a solution, he reduces the concept to "behavioral uncertainty", ignoring environmental uncertainty [@watjatrakul2005 391-392]. This is in line with the original interpretation by @williamson1985 [79]: "The proposed match of governance structures with transactions considers only two of the three dimensions for describing transactions: asset specificity and frequency." Uncertainty arises from these two, and together with bounded rationality and opportunism, gives rise to exchange difficulties [@williamson1975 7], making it "more imperative to organize transactions within governance structures that have the capacity to 'work things out.'" [@williamson1985 79] In this context, behavioral uncertainty can't be disentangled from asset specificity. @williamson1985 [78] is fully aware that this is a departure from Coase's transaction cost rationale.

To drive back the theory to the subject of consulting, @canback1998 [37] argues that it's low mainly human asset specificity that favors the use of consultants, referring to their solutions and advice that can easily be reproduced at many organizations.[^4] @watjatrakul2005 [408] put the theory to the test and compared the transaction cost view with the resource-based view (infra) for describing the sourcing decisions in three cases and comes to the following conclusion: "a high-specificity asset has a major impact on sourcing decisions. It overpowers the effect of uncertainty."

[^4]: Furthermore, Canback claims that transaction frequency and uncertainty are less of an influence. By referring to market uncertainty, not only does he obscure the fact that consultants rather thrive in a context with high complexity and uncertainty, he also misrepresents the uncertainty dimension that is central in transaction cost economics. This is a prime example of the vagueness surrounding the concept of uncertainty in transaction cost economics (supra).

Focusing on low-specificity assets allows consultancy firms to achieve economies of scale. That's why they rather shun highly idiosyncratic assignments. Rather, they'll focus on (often high-level) organizational advice and IT architectures, since these have the biggest adaptive properties.

Borrowing rhetoric from the resource-based view (infra), @mata1995 [498] applies Canback's conclusion on technical IT skills: "While technical skills are essential in the use and application of IT, they are usually not sources of sustained competitive advantage. [...] they are usually not heterogeneously distributed across firms. Moreover, even when they are heterogeneously distributed across firms, they are typically highly mobile. [...] firms without the required analysis, design, and programming skills required to make an IT investment can hire technical consultants and contractors." Ergo, digital capabilities are very likely to be outsourced.

@nevo2007 [16-17] also concludes that his research supports the transaction cost hypothesis: "when the internal IT capability is weak, developing and implementing an IT solution is likely to cost more compared with hiring external IT consultants to do the same job."[^5]

[^5]: The reverse situation also supports the identification theory: "IT consultants will not receive the legitimacy they require [...] if their knowledge and expertise do not differ from that possessed by the in-house IT team. Under these circumstances, external IT consultants' impact on IT productivity is expected to be lower." [@nevo2007 17]

## Agency Theory

At the center of agency theory is the observation that the firm is not a single individual. *Au contraire,* "the *behavior* of the firm is like the behavior of a market; i.e., the outcome of a complex equilibrium process." [@jensen1976 311] When (the owners of) an organization ask(s) its employees to deliver a service, or buys a service in the market, it encounters the two-pronged principal-agent problem[^6]. By delegating work to an agent, the principal has to account for the fact that "(a) the desires or goals of the principal and the agent conflict and (b) it is difficult or expensive for the principal to verify what the agent is doing." [@eisenhardt1989 58] These phenomena are known as goal incompatibility and information asymmetry.

[^6]: Agency theory is also concerned with the problem of risk sharing that arises when principal and agent each prefer different courses of action due to differing attitudes towards risk. [@eisenhardt1989 58] However, this is not within scope of this paper.

-   *Goal incompatibility*: The organization is interested in a timely roll-out of a quality solution for a problem they have. The consultancy firm, on the other hand, is driven by profit maximization.
-   *Information asymmetry*: consultants wield enormous power over the knowledge that they possess and use. Their clients depend on this knowledge, making them vulnerable, and putting them at mercy of the consultant. [@brien1998] For example, @ylonen2019 [248-249] found that payments from the Finnish Ministry of Finance to a particular consultancy company drastically increased year-over-year, because the consultancy firm had a quasi-monopoly in the knowledge regarding particular remuneration practices. Furthermore, when a particular individual sells his services to another one, it may be difficult to asses its true value [@ouchi1980 134-135]. This is especially the case when interdependent technologies are involved, as their implementation and maintenance requires teamwork. On that account, disentangling individual contribution from the team's joint efforts are particularly hard. @aubert1996 [59] also points to a problem of measurement within IT outsourcing. Contracts often specify all kinds of measures: response time, uptime, error logs, etc. Although they are linked to explicit provisions (such as fines, penalties and contract termination), there are two conditions for them to be effective: (1) observability and (2) verifiability. The former implies that the client can observe the actual performance of the agent, while the latter is about verifying observations and providing evidence. This situation invites opportunism such as slacking off.

These phenomena could lead to opportunism such as *adverse selection* on the one hand and *moral hazard* on the other hand. [@rousseau1993 14] The former refers to the contracting of agents "unqualified to fulfill their end of the bargain due to active misrepresentation of the agent's competence and expertise." The latter occurs when "the agent shirks and reduces his or her efforts."

To prevent adverse selection and/or moral hazard, control mechanisms could be put into place, resulting in agency costs. These include the costs of structuring, monitoring, and bonding a set of contracts among agents with conflicting interests, plus the residual loss incurred because the cost of full enforcement of contracts exceeds the benefits." [@fama1983 327] The following expenditures sum up to the total amount of agency costs [@jensen1976 308]:

1.  monitoring expenditures;
2.  bonding expenditures to achieve incentive alignment;
3.  residual loss: the remaining "loss of welfare" of the principal in those situations where the agent makes divergent decisions because it was to expensive to offset it through bonding or monitoring.

Circling back to the this paper's subject of digital consulting, we can summarize the interplay of transaction cost economics and agency theory as follows. Low asset specificity of knowledge production and the improvement of internal coordination mechanisms through IT leads firms to order these services in the market. However, human bounded rationality does not allow for detailed consultancy contracts. Payment for consulting engagements are typically recurring (high frequency), which leads to behavioral uncertainty, since the incentives of the consulting firm and the client do not align (goal incompatibility), their complex interdependent services are hard to evaluate and their contributions are hard to measure (information asymmetry). This could lead to adverse selection and moral hazard, and asks for control mechanisms to economize on agency costs.

Finally, an extra remark is in place. One should not forget that managers are also agents themselves. As @tosi1997 [584] describes: "the reality [is] that in large organizations, owners may be separated from the managers who make decisions in forms, and that the two may have different interests." In that sense, @fincham2002 claims that a consultant can be described as "an agent's agent", extending the management's own agency function. In other words, consultants operate at the "outer reaches of corporate power", stretching corporate authority to its limits. Consequently, their legitimacy is often problematic within the corporation that engages with them.

## Resource-based View

The resource-based view rejects the traditional micro-economic assumptions that resources are homogeneously distributed across firms [@barney1991 99-103]. Instead, it argues that they are heterogeneously distributed, which could allow a firm not only to establish a competitive advantage[^7], but also a sustained competitive advantage when these resources are not perfectly transferable, for example because of resource immobility. These resources come in the form of assets, capabilities or organizational processes. Firms can obtain above-normal results if they can establish a competitive advantage by making their resources exploit opportunities in the market, or neutralize those established by competitors. To be strategic, resources should be valuable, rare, inimitable and non-substitutable.

[^7]: "A firm is said to have a competitive advantage when it is implementing a value creating strategy not simultaneously being implemented by any current or potential competitors." [@barney1991 102]

@mata1995 [495-500] identifies five attributes of information technology as possible sources of sustained competitive advantage for a firm.

1.  Access to capital: IT (or digital) investments can be very expensive and risky. Only a couple of firms might be able to acquire the capital to make these investments, making it a source of sustained competitive advantage.
2.  Proprietary technology: when technology can be kept proprietary, it can be a source of sustained competitive advantage. However, the author states that one the one hand, it's very hard to prevent proprietary IT knowledge from leaking, and on the other hand, IT has become generic more generic. We'd like to add that many organizations' digital ecosystems are peppered with open-source technologies[^8], which is freely available. In other words, for most non-software firms, a sustained competitive advantage through proprietary software is unattainable.
3.  Technical IT skills: the ability to attract and keep technical IT skills that is required to built IT applications. However, these skills are rarely immobile, nor are they really heterogeneously distributed, according to the author.
4.  Managerial IT skills: management's ability to conceive of, develop and exploit IT applications to support and enhance other business functions can enable a firm to manage market risks associated with IT investments. Given the entanglement of business and IT skills, one could say that managerial IT skills are heterogeneously distributed, and that they are immobile.
5.  Customer switching costs: the ability to create a lock-in effect for one's customers through the use of IT. The author states that switching costs, in many situations, are not a source of sustained competitive advantage, as customers will anticipate it, lock-ins create reputation damage and there are many digital alternatives available. However, there are various value-creating attributes that can be achieved through a lock-in. For example, @amit2001 [505-507] describe how customization, two-sided markets, trust-establishing rating systems or virtual communities are features that generated customer switching costs yet are also value generators deemed legitimate.

[^8]: For example, a survey by the European Commission found that around 40% of respondents stated that open-source is the most important source of software development in their organization [@ec2021 191].

@willcocks2003 [188-189] describes how sourcing decision related to technical IT skills depends on both the market for IT consultancy skills and the type of business activity.

-   By market comparison: A high-cost, low-quality market leads to in-house development, while a high-cost, high-quality market should lead to insourcing. A low-cost, low-quality market leads to cheap-sourcing[^9] and a low-cost, high-quality market is perfect for outsourcing.
-   By business activity: non-critical, commoditized applications should be out-sourced. Critical, commoditized applications should be insourced or built in-house, and differentiating, critical applications should be built in-house or acquired.

[^9]: Cheap-sourcing involves low investments and effort, but also comes with no internal learning. For example: development of a new promotional website by a digital agency.

By referring to commoditization, one could say that @willcocks2003 implicitly refers to asset specificity, blending elements of transaction cost economics in the resource based view. @watjatrakul2005 [405-407] makes this explicit by juxtaposing the resource-based view with the transaction cost view. Four types of assets result from this exercise:

1.  Low specificity, non-strategic such as generic managerial capabilities.
2.  Low specificity, strategic such as a configuration of capabilities that result in certain strategic decisions.
3.  High specificity, non-strategic such as an consumer tracking technology that provides valuable insights into the organization's processes.
4.  High specificity, strategic such as company experts that are responsible for developing a organization's differentiating features.

## Social Identity Theory

The research by @schwarz2005 [311-313] claims that it matters *who* implements an IT project. Given the interdependent nature of consulting, social dynamics are relevant when teams need to cooperate closely with consultants: "technology-enabled inertia can be explained through understanding an employee's social identifications and his or her associated cognitions, where inertia exists on a sliding scale of change." By defending their self-image, low-status groups can hinder the implementation of an application. For example, early studies on social identity theory already showed that "subjects do sacrifice absolute levels of group and personal gain to achieve intergroup differences in outcomes favouring the ingroup." [@turner1979 200]

This theory amends transaction costs economics and the resource-based view by pointing to possible sub-optimal outcomes due to conflicts between in-house personal and consultants. For example, doubling a team's capacity by adding consultants doesn't necessarily produce double the output. Sourcing decisions should account for this non-linearity.

Furthermore, interdependency of consultants with the rest of the team complicate assessing individual contributions. Using social identity theory insights, one could even argue that the in-group's striving for intergroup differences further impedes the control and evaluation of individual consultants.

## Embeddedness Theory

Embeddedness theorists distance themselves from transaction costs economics [@armbruster2006 14-16]. They argue that outsourcing decisions are the byproduct of the relationships between the decision-makers across different companies. Although transaction cost economists stress the importance of trust within relational contracting, it is "undersocialized" [@granovetter1985] according to embeddedness theorists. "As a result, transactions may be inefficient without the participants either noticing or calculating it as such. A transaction cost analysis of such processes may then represent an ex post rationalization of an otherwise inefficient solution." [@armbruster2006 15]

@nooteboom1996 [992] states that typically, economists tend to neglect intrinsic utility and that it doesn't matter who the transaction partner is. Embeddedness theory rejects this, as personality and social embeddedness enter the picture. The exchange process an sich matters as people may prefer to "transact on the basis of trust and its concomitants of ethics, kinship, friendship or empathy."

In the language of transaction cost economics and signaling theory: embeddedness would be a tool for preventing adverse selection as "[p]rincipals frequently know their agent's type because of personal familiarity with potential agents or through members of trusted social networks in which both principal and agent are embedded." [@shapiro2005 277] Furthermore, it also prevents moral hazard as "[n]orms and values of conduct may mitigate inclinations towards opportunism; they may cause people to forego opportunities for opportunism, at least within reasonable bounds." [@nooteboom1996 990]

Several empirical findings support embeddedness theory. For example, @kitay2004 [15] found that "in many instances consultants and clients develop closer, on-going relations in which the organisational boundaries between the parties become blurred (an 'insider' role)." However, this is not always the case. The authors note that there are "relationship firms" and "transactional firms." According to them, "criticism directed at management consultants could simply be instances where client and consultant perceptions diverge in terms of the type of relationship that each party is seeking."

## Sociological Neoinstitutionalism

A theory that is systematically drawn upon [@armbruster2006 6-8] is sociological neoinstitutionalism. It is very critical of transaction cost economics, and further complicates the issue of adverse selection. Its central argument is that the belief in efficiency of certain practices or solutions drives economic actions, rather than the proven efficiency.

@jayatilaka2006 [120] argue that "[i]nitiatives to follow other companies in IS sourcing arrangements could originate from IS and managerial professionals working in the companies and consultants who are aware of successful outsourcing arrangements. Mimetic isomorphism occurs when companies follow the lead of other companies that have successfully outsourced IT."

For example, @loh1992 [350] argues that the decision to outsource their entire IT department to IBM by Eastman Kodak in the late 1980s had a tremendous impact on other companies' outsourcing decisions.

Mimetic isomorphism is driven by three mechanisms [@dimaggio1983 150-154]:

1.  A coercive authority formally or informally exerts pressure on organizations to collude or fall in line. This authority can be governmental (through laws & regulation) and non-governmental (through standard operating procedures established at the conglomerate-level).
2.  Uncertainty regarding technology or organizational processes could drive organizations to "model themselves after similar organizations in their field that they perceive to be more legitimate or successful. The ubiquity of certain kinds of structural arrangements can more likely be credited to the universality of mimetic processes than to any concrete evidence that the adopted models enhance efficiency."
3.  Normative pressures mainly stem from professionalization (infra), as the members of an occupation try to establish a "cognitive base and legitimation for their occupational autonomy."

The result is that large consultancies have been described as carriers, not only of knowledge, but of legitimacy too. After all, it's their analyses that validate management decisions. @mazzucato2023 [16] describes consultancy as confidence tricks "in contracts with hollowed-out and timid governments and shareholder value-maximizing forms."

@mazzucato2023 [21] states that "[t]he consulting industry often provides legitimacy for controversial decisions. When a corporate senior manager wishes to convince their board of something, or when a government minister wants to win over others to their vision or stall meaningful action, a supportive report from a Big Three or Big Four firm can go a long way at the expense of other objectives--or even labor agreements.

@kieser2006 [69] agrees: "Companies that are held internally and externally accountable for how they 'handle' uncertainty will contract consultants as a sign of good management."

@zucker1985 [20-21] argues that this is the result of trust-building signals (infra) growing beyond their initial goal of delineating specific expectations. Trust-producing firms (such as consultancy firms) an sich assume a high status, with the business world protecting them against failure. This entrenchment acts as a flywheel for the growth of the consultancy market as a whole.

To a large degree, the critical work by @mazzucato2023 [120] is inspired by sociological neoinstitutionalism: "The Big Con does not just describe discursive tricks, but also how the practices of the consulting industry--what we call"consultology"--combine with the broader structures of our political economy to extract rents from clients, often by enabling those clients to extract rents themselves. Ultimately, this entrenches businesses' and governments' reliance on consultancies."

## Signaling Theory

Another theory that falls in the camp of the critical view is that of economic signaling theory [@armbruster2006 8-10]. Unlike sociological institutionalism, it treats the economic actors as experienced and knowledgeable, and not as part of an institution. Signaling theory focuses on the deliberativeness of individual consultants and firms by investing in features that signal status, quality and reliability. This is one of the main reasons why top consultancy firms invest so much in attracting top students from renowned educational institutions: "in a market of experience and credence goods, the quality of human resources signals consulting quality." [@armbruster2006 178]

Especially in a situation where past transactions are absent, "management consultancies must convey in some way to their clients that they have something valuable to offer. [...] consultants are able to take control of the process by which impressions and perceptions of their service are created. By managing the creation of these images, consultants are able to persuade clients of their value and quality. Management consultancies are therefore 'systems of persuasion' *par excellence* and impression management is not external to the core of their work but is at its core." [@clark1998 35]

Although @leicht2006 [35] do not deny a changing economical and institutional context, they also observe the opportunism of consultancy firms. "Management consulting is the new and relatively recent attempt to take advantage of the destructured business environment of corporate clients that purchase business services. They represent a semi-institutionalized attempt to advance the professional aspirations of managers themselves, especially in light of the well-publicized attacks on middle management infrastructure that has accompanied the latest corporate downsizing waves."

@gill1993 [290] notes that for managers, "[t]erms such as 'growth' and 'effectiveness' have mythical qualities. They are condensation symbols collapsing a managerial world view into a single word. So, too, [...] consultancy packages [make] use of condensation symbols thereby creating affective bonds to the symbol's object, tying managers into the package at an emotional level and creating a shared managerial language." @zucker1985 [15-16] agrees on the role of similarity. While nationality, ethnicity and sex can indicate a common cultural system, or a "world held in common", more superficial (bought or acquired) features can delineate specific expectations in specific situations. In consulting this translates into degrees, certificates, using the adequate buzzwords, wearing a suit and driving a quality car. These indicators signal adherence to the "rules of the game."

Sociological neoinstitutionalism argues that legitimacy-seeking behavior leads to inefficient market outcomes, while signaling theory argues the opposite.

# Problem statement

## Types of Opportunism

Organizations employing the services of a contingent workforce, like digital consultants, should always be aware of opportunism arising from information asymmetries and goal incompatibilities, as outlined in agency theory. Two types of opportunism can be identified @clark1993 [242]:

1.  Ex ante: Pre-contractual opportunism or adverse selection
2.  Ex post: Post-contractual opportunism or moral hazard

### Ex Ante Opportunism: Adverse Selection

Adverse selection is associated with the client's inability to determine the client's capabilities with regards to the assignment. This analogous to Akerlof's "Lemons problem" [-@akerlof1970]: due to the information asymmetry, clients don't want to pay more than the average price for consultants within a certain niche. While consultants of below-average quality ("lemons") benefit from this average price, above-average consultants will not want to compete, and are crowded out. Furthermore, the economic barriers to entry [@fee2004 463] in IT consultancy (and consultancy in general) are few to none. Anyone with experience in a specific field, sector or technology can wrap it as advice and sell it to whoever wants to hear it.

Typically, in this situation, a sector tends to professionalize, or become regulated. However, in (digital) consultancy, this is not the case.

#### No Professionalization

Professionalization is a mechanism that typically protects clients against self-proclaimed experts but who are unqualified and expose their clients to even greater risks than the ones they are supposed to advice on. [@kieser2006 71] Furthermore, they also protect experts against colleagues that could ruin the reputation of the whole industry. Finally, professionalization typically results in the establishment of an ethos, which fosters[^10] trust between a profession and the public at large [@sokolowski1991] Through these mechanisms, "professions are social devices to limit agency costs." [@shapiro2005 276]

[^10]: A professional ethos is "a set of written and unwritten rules that guide professional practice." [@enstad2017]

Certain countries, like Germany, have made attempts to professionalize the sector through the establishment of specialized university programs to make the use of the title "consultant" dependent on the attainment of specific qualifications. These attempts haven't turned out to be successful: today, consultancy is not yet a profession, not in the classical sense. [@kieser2006 73].

According to a popular functionalist view [@goode1957], which identifies criteria that distinguish professions from occupations: (1) members share a common identity, (2) it's a life-long calling, (3) share common ideals, (4) have a shared self-conception, (5) behave distinctly towards non-members, (6) use specific language, (7) easy to recognize and (8) follow an ethical code by a powerful association.

The power approach, on the other hand, is much different from the functionalist approach. It argues that professions are a group of people that wants to achieve expert status within society, attaining certain privileges. [@kieser2006 75]

Both are potent frameworks for providing arguments why the practice of consultancy is not likely to professionalize. From the functionalist perspective, we identify the following:

-   Consultants do not seek to differentiate themselves from their clients, *au contraire*, they seek partnership, as to truly understand the problems and opportunities a client faces. [@fincham2006]

-   A large part of the *raison d'tre* of consultants relates to their capability of transferring expertise to the clients. This differs from professionals, who are approached by clients to solve a problem, without the expectation of knowledge transfer. [@oakley1993]

-   Consulting organizations are extremely diverse, both in terms of subfields (IT, HR, strategy, ...) and in terms of organization size, structure and business models. [@kieser2006 79, 89]

The arguments of the power approach are the following:

-   On the one hand, there are the small consultancies which could try to establish associations to build up their reputation. On the other hand, there are the big consultancies, with a high reputation, that only stand to lose because of increased competition with (a priori) lower reputation organizations. For this reason, the probability of establishing all-encompassing associations is rather low [@kieser2006 77].

-   Many employees of large consultancy firms see their tenure as a platform to self-employed consultant or partnership at a smaller consultant. A title, or membership of an association would facilitate further split-offs at large consultancies [@kieser2006 80].

One could even argue that professionalization is not even relevant in today's capitalism. The "individual given class status, autonomy, social elevation, in return for safeguarding our well-being and applying their professional judgement on the basis of a benign moral or cultural code [...] no longer exists." [@dent2013, 1-2] Trust and respect in today's professional is "earned through their ability to perform to an externally given set of performance indicators." In that sense, "consultants are the real champions of creating a new professional *appearance*." [@kieser2006 95] By combining a promise of market success and professionalism they are "market professionals".

#### No Regulation (TODO)

-   Three sources of regulation can be identified (@clark1993 246-247).
-   See @muzio2011 [813-817].

While in some countries, like Austria, there is a registration system for consultants, governments have been ambivalent towards professional regulation of the consulting sector [@muzio2011 814]. According to @leicht2006, this is the result of (Western) governments embrace of market logic and their reluctance to intervene with "additional coercive pressures in the wake of questionable professional conduct."

One important antecedent for regulation is professionalization of the sector, which is highly unlikely in the current institutional environment. Although "expertise is increasingly more important in differentiating societies, it is difficult to defend the status of professions, to say nothing of establishing new ones. The trend is towards deregulation and deprofessionalization of existing professions, rather than towards establishing new ones." [@kieser2006 90]

Finally, governments are often large consumers of consultancy services, making them wary of supporting regulation of the sector, as it could drive up prices and reduce their autonomy in hiring them [@muzio2011 815].

In the existing literature, these remarks are part of a critical paradigm regarding consultants [@armbruster2006 4-5]. Authors point to the contestable nature of consulting, the self-interest of consultancy firms, and the stretching of consultancy advice.

See also:

-   [@wright2002]
-   [@david2013]
-   [@mahoney2016]

See @sturdy2021 [3-4]

See @sturdy2021 [4]

### Ex Post Opportunism: Moral Hazard

Moral hazard is the result of two elements at the core of transaction cost economics and agency theory: the ambiguity of the measurement of individual performance and the goal incompatibility between principal and agent during the cooperation. Ergo, when something goes awry, it is challenging for the client to convincingly pinpoint the blame on the consultant, due to their interactive nature and embeddedness in the organization.

@mcfarlan1995 [9] also points to a relevant aspect. Because the target state of a project might change, most engagements are embedded in a flexible contract, exacerbating the difficulty of blaming the consultant: "[I]n each subsequent year, the contract payment stream becomes less and less tied to the initial set of planned outputs (as the world changes) and, thus, more subject to negotiation and misunderstanding." @nooteboom1996 [924] agrees, stating that "detailed contracts can shroud a contract in an atmosphere of mistrust, derailing the relation before it has even started."[^11]

[^11]: At first sight, the work by @lacity2012 [4] seems contradictory to these findings, as it describes that there is substantial evidence that positive outsourcing outcomes are associated with *more detailed contracts*, shorter-term contracts and high-value contracts. However, these details relate to scope, service levels, responsibilities and adaption to change, but not necessarily to the individual tasks that comprise the assignment.

For @mazzucato2023 [145], the resulting skewed risk-reward relationship is at the heart of the consulting industry's business model: "The rewards reaped--the rents--usually far exceed the financial risks of taking on the contract or the costs of creating an impression of value [because] large consultancies today can survive knocks to their reputations in the wake of public scandals with the help of their extensive resources, employing savvy legal and PR teams to respond to newspaper exposs and government inquiries."

The ways that moral hazard can occur between a client and its consultants are plenty.

-   Shirking due to individual consultants slacking off or shifting attention to other clients.

-   Locking in a client by using proprietary tools, frameworks and technologies that are not widely supported.

-   Selling the initial contract cheap, but the services that follow come with a premium. This practice of "making an offer more than possible to deliver and revoking a central part of the offer after the target subject accepts" [@motes1979 219] is known as "lowballing".

-   Mishandling of confidential information, which not only includes straightforward transfer of information to competitors. Information may also be benchmarked with data from other firms without a principal's knowing. Furthermore, project reports, presentations, and analyses are saved in internal knowledge databases and may be used when similar projects come up. [@armbruster2006 72-73]

It's clear that contractual flexibility is a double-edged sword. On the one hand, it positively impacts the diversity of tasks in which managers can employ consultants, enabling them to respond quickly to threats or opportunities. On the other hand, it decreases consultants' formal accountability.

## Opportunism in Digital Consultancy

There are three main reasons why "digital consultancy" deserves a dedicated scope

Information technology evolves rapidly. Consequently, organizations risk a high degree of both ex ante and ex post uncertainty. Will consultancy firms be able to deliver in n years from now? Will they customize the solution to such a degree that only they can deliver support, and lock in their client? Especially organizations with a low technological maturity have the highest risk encountering opportunism when working with third-party service provides.

Ambiguity of measuring individual performance. When a particular individual sells his services to another one, it may be difficult to asses its true value [@ouchi1980 134-135]. This is especially the case when interdependent technologies are involved, as their implementation and maintenance requires teamwork. On that account, disentangling individual contribution from the team's joint efforts are particularly hard. This situation invites moral hazard, such as consultants slacking off, inflating their time sheets or unnecessarily widening the scope.

Finally, there is often a degree of lock-in involved as some firms have exclusivity for implementing or personalizing a certain solution. as the costs of switching providers are often significant, it is not straightforward to switch vendors when ex post opportunism is detected.

## Preventing Opportunism

In the existing literature, several governance mechanisms for preventing opportunism have been proposed. In the governance literature, there is a strong focus on control, however, several of these mechanisms aren't meant to exert control, but rely on, or foster trust.

A "control" is defined as "an action to ascertain whether another action has been successfully executed, or if a certain state has been achieved or maintained, in order to deal with possible deviations and unforeseen events in order to positively cope with them." [@castelfranchi2000 809]

On the other hand, the following definition of trust is adhered to: "the willingness of a party to be vulnerable to the actions of another party based on the expectation that the other will perform a particular action important to the trustor, irrespective of the ability to monitor or control that other party." [@kee1970]

By relying on these definitions, the explicit choice is made that "trust in [the agent] and control of [the agent] are antagonistic: where there is trust there is no control, and vice versa; the larger the trust the less room for control, and vice versa. But they are also supplementary: one remedies to the lack of the other; they are parts of one and the same entity." [@castelfranchi2000 813]

The concepts of trust or control (or both) have a significant role in three of the earlier discussed theories. From a transaction cost economics perspective, trust makes transactions cheaper and allows for greater flexibility, as it requires less incentivization, and specification and monitoring of contracts [@nooteboom1996 989]. It also ties into embeddedness theory which claims that strong relational ties between the agent and the principal are what drive decisions to outsource to (and to which) consultants. Finally, signaling theory is about trust-building systems of persuasion.

@liberatore2010 [265] confirms that building trust drives the consultancy firm's motivation for short-term profits towards long-term business and reputation.

However, trust is not static. @kirilov2012 [193-194] breaks trust factors down in trust-building and trust-sustaining factors. The former is about signaling ability through references, experience and reputation. The latter is about signaling integrity: effective and transparent communication, proactivity, monitoring and consistently meeting contractual obligations. In that sense, trust can be describe as a "Bayesian-like decision process" in which all trust-relevant information is carefully scrutinized over time to assess the proportion of "cooperative choices or long-term behavioral patterns." [@lewicki2006 995]

@shapiro1992 [366-374] identified three bases of trust. Investing in either of these can be matched to the level and benefits of the trust desired, and the costs and risks associated with it. The three bases of trust are the following.

1.  Deterrence-based trust -- or "calculus-based trust" [@lewicki1996 119] -- results in predictive behavior of the agent because there are measures in place to prevent hostile actions. These measures generate a potential cost for acting in a distrustful way, outweighing the advantages of doing so.
2.  Knowledge-based trust emerges from prior contacts based on the premise that through ongoing interaction, partners get to know each other better and develop trust around norms of equity [@gulati1995 92]. People act cooperatively when they expect their partner to behave in the same way and reciprocate. Each has something to give the other. Small gestures, such as a consultant providing market insights, a client inviting consultants to team buildings or company events, pressure each other into conformity [@macaulay1963 63].
3.  Identification-based trust rests on the premise that when both parties share the same preferences they tend to behave in a more trustworthy manner towards each other [@shapiro1992 371]. Several conditions can build trust based on identification, of which joint products or goals, a common team name, proximity and shared values are mentioned explicitly in the research. An interesting finding by @schoenherr2015 [408] is that process integration not only serves as a mechanism for information exchange, but also for engendering reciprocity, yielding greater levels of trust.

Like @kirilov2012, @lewicki2006 [1011] describes that the bases for trust change over time. For example, as parties get to know each other, their trust evolves from deterrence-based to knowledge-based. However, rarely does a transition occur from knowledge-based trust to identification-based trust, as it requires identification with each other and the development of strong affect between two parties.

To conclude this section, we note that control and trust are antagonistic, but are both tools for managing agency phenomena, resulting in confidence[^12] that the agent is competent, and willing to do what the principal delegates to them [@castelfranchi2000 802-813].

[^12]: The concept of "confidence" is defined as "the degree of belief in a given hypothesis." [@tversky1992 412]

Below, a fairly exhaustive list of proposed governance measures is outlined and grouped by their corresponding trust antecedent.

| Control Mechanism      | Description | Opportunism       | Agency phenomenon     |
|-------------------|------------------|------------------|------------------|
| Public Reputation      |             | Adverse Selection | Goal Incompatibility  |
| Experience-based Trust |             | Adverse Selection | Goal Incompatibility  |
| Networked Reputation   |             | Adverse Selection | Goal Incompatibility  |
| Procurement            |             | Adverse Selection | Goal Incompatibility  |
| Third-party Assessment |             | Adverse Selection | Goal Incompatibility  |
| Consulting Database    |             | Adverse Selection | Goal Incompatibility  |
| Monitoring             |             | Moral Hazard      | Information Asymmetry |
| Whistleblowing         |             | Moral Hazard      | Information Asymmetry |
| Incentivization        |             | Moral Hazard      | Goal Incompatibility  |
| Third-Party Moderation |             | Moral Hazard      | Goal Incompatibility  |
| First-Party Assessment |             | Moral Hazard      | Goal Incompatibility  |
| Relational Mechanisms  |             | Moral Hazard      | Goal Incompatibility  |

### Preventing Adverse Selection

Several researchers have theorized and/or empirically tested several governance mechanisms for preventing adverse selection of consultancy service vendors.

#### Procurement

"Management consulting is a highly interactive service in which interpersonal trust and 'liking' play a central role. [It is often described] as very similar to recruiting an employee." [@furusten2005 185] Experienced managers buy the services of individuals in whom they have confidence, not from consulting companies. The role if interpersonal is motivated by the argument that the need for consultants is often not driven by the organization, but by the manager's personal needs and insecurities. This ties in to the "agent's agent" argument in @fincham2002, discussed earlier. This leads to the conclusion that managers do not always behave in the company's interest when dealing with consultants, and governance measures are required [@honer2006 300].

Both @sturdy2021 [4-5] and @lonsdale2017 [164] juxtaposes purchase regulation (often embodied in a procurement department) with relational mechanisms (infra). The traditional approach is described as *collaborative* or *relational* contracting. The approach that involves procurement is described as *transactional* or *openly competitive*.

@honer2006 [307] describes how a central purchasing/project office which selects consultants on behalf of the managers could streamline the hiring process of consultants. Furthermore, this office could be tasked with control and coordination of all consulting projects, going beyond mere selection. Nevertheless, this could be incompatible with company culture and managerial budget responsibility. In response, the research proposes the creation of standardized processes for dealing with consultants. These would give managers "clear instructions for dealing with consultancy and enhance the principal's control." However, the author highlights that this practice, too, might impact a manager's perception of their autonomy and they might act in ways to ignore or bypass these rules.

Typically, the involvement of procurement intermediaries commodify[^13] management knowledge [@omahoney2013 205-206]. This could be troublesome, as commodification is often resisted on two fronts, both representing a power struggle.

[^13]: Commodification is "the process whereby an object (in the widest sense of the term, meaning a thing, an idea, a creature, etc.) comes to be provided through, and/or represented in terms of, a market transaction", crystalizing their value into a price. [@carvalho2008]]

-   On the one hand, the "producers" of management knowledge, (i.e. the consultancy firms) are well aware that commodification could potentially neglect their competitive advantages, having a deleterious effect on their profit margins.

-   On the other hand, as procurers define the problem of management, the come in direct competition with the managers they represent.

The latter is confirmed by many of the respondents in @lonsdale2017 [161-163]. They state that involvement of procurement can be a recipe for disaster if they do not understand the project or have no experience with the services that are being bought in the market. This can lead to two adverse affects: (1) selection of a vendor that cannot deliver on his promises and (2) a focus on price, resulting in "ending up with the B team."

@honer2006 [305-306] found that friction between managers and procurement can result in three behavioral patterns:

-   departmentalism: managers claiming that their department, contrary to the company as a whole, is very transparent regarding their use of consultants.

-   authority protection: managers claiming that intervention ins their authority is neither necessary, nor desired and that they have authority to how their department's problems are solves.

-   laziness: managers do not recognize that using consultants effectively is a significant managerial task, and underreport unsuccessful projects.

Nevertheless, @lonsdale2017 [164] claims that, regarding consultancy services, the impact of purchase regulation on service quality is ambiguous, and largely depends on the purchasing phase in which the procurement department is involved: "In terms of the more positive perceptions of involvement at the needs assessment stage, as against the sourcing and contract management stages, this was seen as due to procurement involvement here posing less of a threat to end-user interests and requiring lower levels of service-specific knowledge."

#### Third-Party Assessments

@zucker1985 [57-62] describes the rise of the "social overhead sector" in the 19th and 20th century. This sector acts as an "intermediary" or a "broker" in a variety of situations where transactions need to be bridged: stock brokers, real estate agents, banks, etc. These firms, bureaucracies and individuals are specialized in the production of trust, and marketing it to others. The same principle can be applied to the consultants: assessment by a third-party agency to prevent adverse selection.

However; according to some, assessing the quality of consultants is impossible. For example, according do @bloomfield1995 [40], "there can be no presumed separation between technical skills and political skills, and no ranking between the two in terms of their importance for consultancy practice and the development of IT in user organizations." Furthermore, @bettencourt2002 [101-102] states that for knowledge-intensive business services (KIBS) to succeed, a lot depends on the client. "Client co-production roles [...] are emergent, multi-faceted, and highly collaborative because clients themselves possess much of the knowledge and competence that a KIBS firm needs to successfully deliver its service solution."

Another valid point is raised by @basu2011 [23] , who discovered that pre-qualification efforts displayed little effect on adverse selection because the firm might present an excellent reputation, but individual consultants assigned to a project might lack the required skills.

For these reasons, @clark1993 [250], states that one has to rely on informal mechanisms: "the main trust-producing mechanism [...] is the 'closed' social structure; a form of individual trust. The formal, institutional-based, trust-producing mechanisms are weak. It is the contractual guarantees, and the history of past transactions underlying reputation, which overcome the potential effects of adverse selection and moral hazard."

#### Reputation

@kirilov2012 [193] describes reputation as "a mixture of the brand name of the enterprise, executive management background, maturity level, customer references, and independent quality assessments."

@armbruster2006 [75-76] distinguishes three types of reputation:

1.  Public reputation is the perception of a consulting firm's (or individual consultant) past performance or potential. For this reason, large consultancies make it known that they spend a lot of resources on the search and selection of recruits. [@kieser2006 91] While there are few to no barriers to enter the market as a whole with a newly-found consulting firm, public reputation is a huge barrier to reaching its upper end. Public reputation is like a public good <!-- citation needed -->; the information is non-excludable and non-rivalrous.
2.  Experience-based trust relates to personal experience with a specific partner. A positive relation drives future action because a partner is less likely to act distrustfully in one transaction if future benefits are jeopardized. [@shapiro1992 367] However, trust evolves slowly, and maintaining it requires commitment. That's why it is often constrained to a small group of business partners.
3.  Networked reputation is a firm or consultant's reputation within a network of business relations [@gluckler2003 271]. For example, "if a trusted party cannot provide the resources that are needed, their relations can be used in order to obtain trustworthy information about parties one is not connected to." [@gluckler2003 280] Not only consulting partners can make recommendations; @honer2006 [308] proposes building informal networks that would allow managers to communicate their experience in consulting to each other.

Public reputation has a high market scope, as all potential partners are active in a specific market, but it generally results in less certainty, since public reputation is easily manipulated. Experience-based trust produces high certainty, but a low market scope, since rarely, one has had interpersonal experience with all potential partners. Finally, networked reputation sits somewhere in between: via a social network, one can estimate the reputation of various consulting partners and it also produces high certainty, since relations are at stake.

@clark1993 [243-244] asked 55 respondents about the factors that are important when choosing an executive search & selection consultancy. "Reputation of individual consultants" and "reputation of the consultancy" are in the top three factors. This reputation often arises from "a history of past transactions with individual consultants. Frequent transactions between consultants and clients leads to familiarity which underpins the latters' assessment of the former." In other words, because finding a new consultants implies a search cost [@wilson2012 1072], incumbent consultants are expected to receive new contracts as long as the cost incurred from a potential sub-optimal performance is lower than the search cost of finding a new consultant.

The findings in @clark1993 are confirmed by @richter2009 [285] who found, with regards to projects that involve client-specific information, that "clients are willing to involve external consultants with whom they have established a relationship of trust in the execution of such projects. [...] intermediate forms of governance between the extremes of market procurement from an anonymous provider and fully-fledged integration, are not only viable, but an effective option for clients to procure managerial services." Organizations are more inclined to work with consultants with whom they have no experience when no client-specific information or industry expertise is required. A prior client/supplier working relationship is also listed as an important determinant in the meta-analysis by @lacity2011 [235].

According to @nayyar1990 [516] "reputation performs as an implicit contract. It is enforced by the seller's concerns about future demand for the service provided. [...] reputation is likely to exhibit characteristics of a public good. Once acquired, it can be user over and over again in the context of other services or markets."

According to a survey with 150 German companies [@kieser2006 91], for 73% of respondents, reputation is a deciding factor. Furthermore, 50% of respondents indicated that recommendations from managers of other companies is a deciding factor.

#### Consulting Database

One way to structure and crowd-source experience-based trust is by centralizing it in a "consulting database". This could force hiring managers to report on their experiences with individual consultants or consultancy firms.

In a study, @mohe2006 set up an *infobase* in DaimlerChrysler that can be used by internal managers for managing consultants and consulting projects. It contains news and trends regarding consulting, process guidelines, consultant profiles, past projects and management info (such as price).

@honer2006 [308], takes it one step further and proposes an institution within an organization that has the staff and capabilities to support managers in dealing with consultancy. This institution would hold data on previous consultancy engagements, assisting managers in choosing the adequate consultancy partner. Their service is entirely voluntary and secures discrete conduct and sensitivity when dealing with confidential information.

### Preventing Moral Hazard

Besides the prevention of adverse selection, several mechanisms have been proposed for preventing moral hazard.

#### Monitoring

[@basu2011 13] states that "Monitoring refers to the application of a feedback system to provide information to the principal about the actions of the agent. [... It] places an uncomfortable social pressure on the agent that increases compliance. It also increases the principal's ability to detect the agent's opportunism and thus to appropriately reward or sanction agent behavior. It reduces the agent's motivation to justify a failed strategy, and promotes actions consistent with shareholder goals." In other words, there are two ways that monitoring affects the relationship with a consultant. On the one hand, the sole act of monitoring pressures the consultant into compliance. On the other hand, it increases the hiring organization to reward or sanction the consultant's behavior.

Researchers have emphasized three broad ways in which consultants can be monitored [@basu2011 15]:

1.  Deliverables: When client organization plans the project and sets measurable outcomes (deliverables), it obtains agreement from the consultancy firm to the specified deliverables and accompanying deadlines.
2.  Planning: the client verifies that the deliverables are being produced according to the original plan by thoroughly and regularly assessing reviews and written and oral progress reports. Meeting with consultants is essential to ensure that consultants share all relevant information in a timely manner.
3.  Quality and scope: during the implementation of a project, the client checks that the consultants do not sacrifice quality nor scope to meet deadlines. Furthermore, the originally committed staff should not be changed without approval.

With regards to monitoring consultants, @shapiro2005 [275-280] makes several interesting remarks:

1.  Duration of a consultancy engagement: monitoring is easier when a relationship is continuous, and is not short-term. This allows the principal to determine a baseline behavior and performance level.
2.  Sharing a social network affords a rich array of sanctions (shaming, ostracizing), increasing the cost of non-compliance for the consultant.
3.  Many regulatory arrangements, such as monitoring, are agency relationships themselves: "Whether they are internal or external auditors, compliance officers, internal affairs departments, government regulators, insurance companies, investment advisors, or rating agencies (e.g., Standard & Poors or Underwriters Laboratory), the monitors are acting on behalf of some set of principals. And, therefore, they too promise agency problems. They shirk, become coopted, engage in corruption, or perhaps simply monitor the wrong things. In an escalating cycle of agents overseeing agents, we must ask: Who monitors the monitors"
4.  Finally, monitoring often relies on proxy measures. Hiring managers monitoring consultants should be aware of the fact that consultants can displace their behavior toward these surrogates in order to appear to be behaving well.

Finally, @shapiro2005 [276] embeddedness (or relational mechanisms) are often "an excuse to relax vigilant recruitment and monitoring, [which] provides cover [...] for confidence swindlers to feign social intimacy and unfettered opportunism." The next section describes several interpretations of relational mechanisms.

#### Relational Mechanisms

One could raise the argument that due to the nature of the services that consultants provice, relational mechanisms trump "hard" governance such as contractual formalities or monitoring. However, according to the research by @lacity2012 [4-5] "contractual governance and relational governance are not substitutes in that a poorly crafted contract cannot be overcome with friendly, communicative, and trusting account managers. Poor contracts, we have found, can make for poor relationships."

Nevertheless, many authors refer to some sorts of relational mechanisms for facilitating the engagement with consultants and preventing opportunism. Below, four constructs, borrowed from one or more disciplines within the humanities: (1) Clans, (2) psychological contracts, (3) cooperative-level contracts and (4) relational contracts.

##### Clans

@ouchi1980 proposes to prevent opportunism through the establishment of a "clan"[^14]. Clans achieve the "union of objectives between individuals which stems from their necessary dependence upon one another. [...] [c]lans display a high degree of discipline [...] achieved through an extreme form of the belief that individual interests are best served by a complete immersion of each individual in the interests of the whole."

[^14]: @ouchi1980 borrows the concept of a clan from @durkheim1997 [127] who gives the term 'clan' to "a horde that has ceased to be independent and has become an element in a more extensive group [...] It is a family in the sense that all the members who go to make it up consider themselves kin to one another [...] The affinities produced by sharing a blood kinship are mainly what keeps them united. What is more, they sustain mutual relationships that might be termed domestic, since these are to be found elsewhere in societies whose family character is undisputed: I mean collective revenge, collective responsibility and, as soon as individual property makes an appearance, mutual heredity."

Clans differ from bureaucracies and markets in that they don't require auditing or evaluation as it takes place "through the kind of subtle reading of signals that is possible among intimate coworkers but which cannot be translated into explicit, verifiable measures." [@ouchi1980 137] For clans to succeed, the following conditions are required: reciprocity, legitimate authority, common values/beliefs and traditions. When these conditions are not met, clans are merely ceremonial and ritualistic, especially when only traditions remain.

Within the context of digital consultancy, reciprocity is achieved in the sense that everybody within a team, both employees and consultants need to have the freedom to call each other out. Furthermore, managers, product owners and lead developers can assume the role of legitimate authority.

However, the conditions of traditions and common values/beliefs are much harder to achieve in this setting. Mayo construed in 1945 that "by reason of external circumstance [...] groups [that] have little opportunity to form, the immediate symptom is labor turnover, absenteeism, and the like." [@mayo1945 111] The solution might be a partial socialization, accompanied by market or bureaucratic mechanisms [@ouchi1980 138], such as the participation in company meetings, to achieve common values and beliefs.

##### Psychological Contract Obligations

According to @ang2004 [357], the legal interpretation of an IT outsourcing contract is too limited. Instead, they claim that the construct of a *psychological contract* is more appropriate for analyzing the relationship between an IT service supplier and customer. The concept of a psychological contract states that contracts are "idiosyncratically perceived and understood by individuals [and] [s]ubjectivity in the contract leads to disagreement between parties." [@rousseau1993 21]

Consequently, the psychological contract not only comprises the legal contract, but also the unwritten promises, interpersonal relations, and the individual interpretations and perceptions. Since consultancy contracts can become extremely complex (with project descriptions going into the ten thousands of words), and the involved parties entangled in multiple ways, these intangible aspects can gain prominence. The research in @ang2004 [369-70] outlines several psychological contract obligations that positively impact the success of an outsourced IT project.

-   On the supplier side: (1) clear authority structures, (2) knowledge transfer by educating the customer, (3) building inter-organizational teams.
-   On the customer side: (1) clear specification of requirements, (2) prompt payment, and (3) project ownership and monitoring.

Psychological contract theory focuses on mutual obligations, and on the individual consultantsnot the organizations that are the parties of the contract.

##### Cooperative-level Contracts

Closely related is the work by @willcockskern [9-13] that makes a distinction between the contractual level and the cooperative level. The contractual level is about payment for the exchange of services and the transfer of assets, information & consultants. The cooperative level involves formal communication mechanisms; personal investments in time, resources & knowledge; mutual goals & objectives and social bonds. The atmosphere surrounding the former is heavily impacted by developments at the latter. A respondent in @willcockskern [9] states that "the contract is a bit like a nuclear deterrent. You need one and you have got to have a framework, but if you've got to use it you are probably in trouble."

##### Relational Contracts

Finally, @rousseau1993 [10-12] describes a continuum between transactional contracts and relational contracts. On one end of the continuum, the former is based on "short-term monetizable agreements with limited involvement of each party in the lives and activities of the other." On the other end, the latter describes "agreements based upon exchanges of both socioeconomical and monetizable elements, duration which is open-ended and often long term, and a high degree of flexibility."

#### Whistleblowing

We can decompose the principle of whistleblowing[^15] into two constructs. On the one hand there is internal whistleblowing. This can be ad-hoc or organized, when there is a formal procedure for reporting misbehavior, such as by consultants. On the other hand, there is external whistleblowing: the inherent risk of reports of misbehavior leaking into the public or even into the press.

[^15]: We define "whistleblowing" as "the disclosure by organization members (former or current) of illegal, immoral or illegitimate practices under the control of their employers, to persons or organizations that may be able to effect action." [@near1985 4]

##### Internal Whistleblowing

In @ayers2005, an experimental setting investigates if two models of ethical decision-making apply to whistleblowing in a scenario of consultant wrongdoing.

1.  Schultz' [@schultz1993] model states that the decision to report relies on three (subjective) factors: the seriousness of the act, the individual's attribution of their own responsibility and the perceived costs of reporting (mainly reprisals).
2.  According to [@reidenbach1990], ethical beliefs aren't unidimensional. To determine the seriousness of a fact, individuals use more than one rationale and different ethical philosophies. In response, the authors developed a multidimensional ethics scale. It contains the following three dimensions: a moral equity dimension (is something fair?), a relativistic dimension (is something fair in relation to the guidelines?) and a contractual dimension (is something deontological?).

When combining these dimensions in a single model, personal costs, personal responsibility and moral equity correlate with the whistleblowing behavior. When an anonymous whistleblowing procedure is introduced, the seriousness of the wrongdoing primes.

##### External Whistleblowing

Consultancy firms always run the risk of of being exposed when one of their consultants is involved in some kind of wrongdoing. For example, the International Consortium of Investigative Journalists [@icij20231] reported that PwC consultants helped setting up financial constructions so that powerful Russians could evade Western sanctions that were imposed after Russia's unprovoked invasion of Ukraine. In another article, the ICIJ describes how tax advisory consultants used confidential information, acquired while consulting for the Australian government, to set up new tax avoidance structures [@icij20232].

Both types of whistblowing could foster good behavior and alignment with the hiring organization's goals. However, the first is more suitable as a formal governance mechanism.

#### Incentivization

Another approach to governing consultants is to align "the preferences of the agent and the principal through an appropriate reward structure [which] helps curb the agent's opportunistic behavior." [@basu2011 13] Both @liberatore2010 [264-266] and @tosi1997 [597] find that a higher goal congruence between the consultancy firm and the client is an enabler of project performance. More importantly, contrary to the design and enforcement of contractual obligations, management by self-interest (i.e. incentivization) "has the advantage that it is cheaper than contracts, is more flexible, and it is in the players' own interest to be seen to comply with agreements." [@nooteboom2000 924]

@basu2011 [15] proposes several actions to align incentives: (1) link payment to completion of the promised deliverables [see also @clark1993 243], (2) sharing of cost savings or overruns with the consultancy firm, (3) incentives and penalties related to timely completion of a project.

However, @noteboom2000 [924] describes how incentivization is not a silver bullet as (1) it requires a need for observation, measurement and monitoring. Furthermore, (2) "[h]ow does one measure and monitor degree of dependence, spillovers, and specificity of investments?" It's hard to quantify compensation for such intangible risks. Finally, (3) it is hard to maintain, since competences and external conditions are constantly changing.

A particular form of incentivization is "hostage-taking": "one-sided ownership of specific assets may be balanced by one-sided hostages going the other way."[@nooteboom2000 924] Hostage-taking can be an adequate governance mechanism when there are no trust-related nor legal enforcement mechanisms. Without it, parties can be discouraged from forming long-term relationships. According to @werner1993 [47-48], "hostages" are used in situations where rational behavior would lead to sub-optimal outcome, in the Paretoian sense. In game-theoretical terms, hostage-taking is used to prevent defecting behavior.

In that sense, incentivization and public reputation (supra) are related to one another as the latter can be the subject of former: "The threat of loss of reputation and resulting loss of business that comes from having the word spread about untrustworthy business practice is generally quite an effective deterrent. For this reason, gossip [...] has been identified as a powerful means of social control." [@shapiro1992 368]

#### Third-Party Moderation

@babin2017 [6-7] describes a single case study in which trust in an outsourcing relationship had eroded over time. Instead of parting ways, it was clear that both parties needed each other. A proactive approach in the form of a third-party facilitated workshop successfully addressed the trust problem. "When both sides feel hurt, and do not trust the other side, it can be difficult to take preliminary steps to begin to repair inter-organizational trust. An objective outsider using a formal approach proved useful to initiating the trust repair activities."

#### First-party assessments

While it is possible to prevent adverse selection through tests, this isn't necessarily the case for moral hazard. For example, @mayo1945 [111] claims that "the belief that the behavior of an individual within the factory can be predicted before employment upon the basis of a laborious and minute examination by tests of his technical and other capacities is mainly, if not wholly, mistaken. Examination of his developed social skills and his general adaptability might give better results."

Examining consultants can be performed in various ways. The main way is to make cooperation contingent upon an interview, like a candidate for a job would. But researching consultants can also be done in more informal matters such as gathering information from (for example) their social media or academic work. This research "is directed at assessing the real compatibility, or 'interpersonal fit,' between partners, as well as assessing the degree to which a potential partner engages in predictable behavior." [@shapiro1992 370]

# Research Design

```{r, echo=FALSE}
options(warn = -1)
suppressPackageStartupMessages(require(data.table))
suppressPackageStartupMessages(require(tidyverse))
suppressPackageStartupMessages(require(ggplot2))
suppressPackageStartupMessages(require(ggthemes))
suppressPackageStartupMessages(require(ggpattern))
suppressPackageStartupMessages(require(knitr))
figure_n <- 0

data <- fread('data.csv')
legend <- fread('legend.csv')
```

## Scope

From this literature study, there are several observations. First of all, it is clear that there is a tremendous amount of theory regarding the *raison d'tre* of consultants, their growth, why organizations hire them, and how they should be controlled. However, empirical research, both quantitative and qualitative, is fairly limited.

The second observation is that very few research takes into account the local or cultural context. Some studies have been performed in the Anglosaxon countries. Thanks to Staffan Furusten, there has been quite some research in the Nordic countries. However, when it comes to research about consultants in Western Europe, very little to none is available. Nevertheless, a country such as Belgium deserves a specific focus for various reasons.

-   Many large organizations have their headquarters in Belgium due to a variety of reasons: the proximity to EU institutions, a chemical and pharmaceutical cluster and decades of tax advantages (such as the notional interest deduction).

-   Belgium has a very high tax wedge, with a range of fringe benefits that are particularly interesting for consultancy firms: for example, company cars and copyright compensation. Furthermore, many people choose to become a freelance consultant, due to their after-tax income being a lot higher when using one of many tax optimization schemes available (e.g. the VVPR-bis regime, or liquidation reserves). For example, between 2015 and 2021, the number of freelancers in the region increased with a staggering 73% [@serv2023].

-   Finally, throughout 2023 and 2024, there's been quite the ruckus due to the extensive use of consultants by ministerial cabinets in both the Flanders and Federal government. For example, the ministry of healthcare spent 43 million on consultancy in less than four years. [@healthcare2023]

The third observation is that most research is about "management" consultants or "IT" consultants. In the case of the latter, there is a strong focus on ERP implementations. No research was found that related to "digital" consultants, as defined earlier. Nevertheless, there are many consultants working in digital marketing, data science, data engineering, ecommerce, etc. Their work often goes unscrutinized due to a variety of reasons.

1.  These consultants often do not participate in typical IT rituals that monitor their performance.
2.  They can easily work remotely, due to the nature of their work, and escape oversight-by-proximity.
3.  Consultants are not subject to HR procedures, and rarely is procurement well-equipped to manage consultants in their firm.
4.  Their hiring managers and the teams they work for are rarely digital-savvy.

The focus of this paper is on the use of digital consultants (supra) working for organizations in Belgium.

## Paradigm

This research starts with the assumption that there is a lot of information asymmetry between consultants, and their hiring organizations and managers. Furthermore, since they are (often) both profit-maximizing forms, their goals rarely align in the short term. Consequently, these hiring organizations expose themselves to a lot of opportunism, both ex post and ex ante.

By starting from this axioma, this study clearly puts itself in the camp of transaction cost economics and agency theory, and research questions and used methodology are laced with it.

## Relevance

## Research Questions

Given the stellar rise of (digital) consultancy, recurring scandals, and implications for how companies and our democracies are run, we believe the following questions deserve the highest priority.

> **RQ1: For which tasks do Belgian organizations rely on digital consultants?**

With thins research question, we want to get a clear picture of the tasks that consultants are involved in. The primary framework for structuring the answer to this question is based on the classification from @turner1982.

> **RQ2: Why do Belgian organizations rely on digital consultants for these tasks?**

A logical followup question to RQ1 wants to uncover *why* organizations use consultants for these tasks, instead of relying on existing staff, or recruiting new personnel to execute the task. This research question does not aim to settle the big theoretical debates between the various theoretical perspectives. Instead, the (amended) classification of practical perspectives from @lacity1994 is used to structure the answer.

> **RQ3: How do Belgian organizations determine the success of a consultancy assignment?**

Not only are we interested in finding out why consultants are hired, but also how their work is evaluated. This is an important precursor to answering the next research questions. Without a definition of 'success', it is impossible to assess if an organization encountered agency problems with consultants.

> **RQ4: Do Belgian organizations experience agency problems regarding the digital consultants?**

Once we have an idea of the definition of success for working with consultants, we would like to know if agency problems stand in the way of achieving success.

> **RQ5a: Which control mechanisms do Belgian organizations have in place with regards to adverse selection?**
>
> **RQ5b: Which control mechanisms do Belgian organizations have in place with regards to moral hazard?**

The final two research questions are the meat of this paper. The answers to these questions are structured in line with the proposed control mechanisms from the literature study (supra). Answering these two questions will provide insight into the current state of governing consultants in Belgium firms and potentially broken down by relevant dimensions.

All questions require an answer of descriptive nature. This paper does not have the ambition to investigate potential correlation or causality between the presence of specific control mechanisms and the success of working with consultants. To answer such questions, lots of quantitative data would need to be collected and analyzed. However, as became clear throughout this research, most organizations do not have a success criterion when working with consultants, let alone an objective criterion. Collecting data from a large-scale survey was simply not feasible.

Nevertheless, these questions could provide the basis for normative research to come up with a prioritized set of control mechanisms and in which context they apply.

## Methodology

### Structured Interviews

To answer these questions, eight structured interviews were done with a variety of decision-makers (regarding the use of digital consultancy services) working for an organization in Belgium. Most questions required a binary answer. Yet, due to the fuzzy nature of the topic, they often acted as conversation starters for followup questions.

The initial goal of this research was to use the findings from these interviews in a structured survey, but as described earlier, this proved to be unfeasible. Structured interviews were the weapon of choice for the following reasons:

-   Many concepts in the interview required an oral clarification, since the definition of various concepts is a topic of discussion an sich. Even the written interview preparation documents did not suffice for bringing clarity. For example: does a "consultant" include freelancers? Does it include people working for digital agencies? What if they're just a project manager?

-   The respondents weren't always well-versed in the principle of corporate governance and control mechanisms. For the respondents to answer the question properly, the proposed mechanisms needed to be translated to their particular situation.

-   For some tasks, the respondents didn't use consultants, but the respondents could often elaborate on the practices in other teams and departments. The same applied to the existence of certain controls.

-   The relational mechanisms are often defined by seemingly small details (e.g. Do consultants wear a visitor badge? Is their lanyard of a different color? Are they invited to company events?) which are nonetheless interesting to determine how an organization relates to its consultants. Some respondents could easily sum up some of these company-specific details, which would be hard to capture in a survey.

### Question Set

-   Q1. (binary) Does your organization hire digital consultants?

-   Q2. (binary) Do you make decisions regarding the hiring of digital consultants?

-   Q3. (multiple response) What tasks do digital consultants fulfill in your organization? (1) providing information, (2) providing a solution, (3) diagnosing a problem, (4) recommendations, (5) implementation, (6) building consensus & commitment, (7) facilitate client learning, (8) improve organizational effectiveness, (open) How does your organization define the success of a consultancy engagement?

-   Q5. (open) What facilitates a successful consultancy engagement? Both with the consult, the consultancy firm and within the organization.

-   Q6. (multiple response): Why do you hire digital consultants? Why do you not recruit internally for these responsibilities? (1) cost reduction, (2) cost control, (3) outside view, (4) facilitating M&A's, (5) missing skills, (6) better performance, (7) knowledge transfer & diffusion, (8) political reasons.

-   Q7. (binary) Has your organization ever hired digital consultants which did not have the required skill set?

-   Q8. (multiple response) What mechanisms does your organization use for reducing adverse selection? (1) reputational checks, (2) procurement, (3) purchase regulation, (4) third-party assessments, (5) consulting database.

-   Q9. (open) Are the controls to prevent adverse selection formal? Are they actively used? Do they work?

-   Q10. (binary) Has your organization ever hired digital consultants that resulted in some kind of abuse of the organization? Examples: using company secrets and insights elsewhere; poaching employees; slacking off?

-   Q11. (multiple response) What mechanisms does your organization use for reducing moral hazard? (1) first-party assessments, (2) monitoring, (3) whistleblowing, (4) incentivization, (5) third-party moderation, (6) relational mechanisms.

-   Q12. (open) Are the controls to prevent moral hazard formal? Are they actively used? Do they work?

## Respondents

The respondents in this interview are senior and mid-level managers from 8 different organizations that have a significant presence in Belgium. Six out of eight were even founded in Belgium. The number of employees working in Belgium ranges between 250 and 3000 (numbers rounded) and excludes consultants. The headcount of each manager *includes* long-term consultants.

All respondents were sampled from the researcher's professional network.

```{r, echo=FALSE, out.width="75%", fig.align="center"}
kable(
  data %>% select(ID, Level, Expertise, Tenure, `Reports to`, N_below, N_Belgium), 
booktabs=T)
```

# Results

Before discussing the results, there are some side notes to be mindful of. \* Multiple respondents are not on the payroll of the company, but work as a freelancer through a management enterprise--which is typical in Belgium. \* Some respondents had a hard time disentangling the concept of disentangling the concepts of a freelancer, temporary worker, agency and consultant, even after receiving a lengthy written briefing. The concept of a consultant was always clarified through six properties from the definition (supra). Many responses required followup questions about the nature of party involved. For example, off-shore workers *can* be consultants, but aren't necessarily so.

## Tasks (RQ1)

As is clear from the answers by the respondents, consultants are hired for a wide range of tasks. However, all respondents indicated that they have hired consultants to provide a solution for a specific problem. Furthermore, six respondents described how consultants were brought in a step earlier: to define and diagnose the problem in the first place.

Also, six respondents described how consultants were used to temporarily bring new knowledge to an organization and "pollinate" in-house employees with it. Respondent 1 stated that for many assignments, client learning is explicitly mentioned in the contract as a deliverable. Respondent 7 indicated that this is rarely a formal requirement from a consultant and knowledge transfer is formally organized through formal training session. The instructors *can* be consultants, but also internal or external experts.

The other tasks were selected five times or less. Task Q3e (implementation) sparked a lot of discussion as multiple respondents indicated that it's company policy to implement as much as possible by people on the payroll. For there is a lot of knowledge produced when implementing a new solution, and they want to reap that externality themselves. Furthermore, they don't want to depend or be locked in with a specific partner. Respondent 4 and 6 agreed, and stated that sometimes "an extra pair of hands" is hired, merely to assist their own people, and this is where the line between a consultant and a temporary worker blurs.

Finally, with regards to Q3d, respondent 8 stated that this is an inherent task of consultants when they work for their organization for a long time. From their outside experience, they should indicate where there is room for improvement.

$~$

```{r, echo=FALSE}
figure_n <- figure_n + 1
data %>% 
  select(all_of('ID') | starts_with("Q3")) %>%
  pivot_longer(cols = starts_with("Q3"), names_to = 'task', values_to = 'in_place') %>%
  left_join(legend, by = join_by(task == Question)) %>%
  mutate(description = paste(task, Description, sep = ': ')) %>%
  select(all_of(c('ID', 'description', 'in_place'))) %>%
  group_by(description) %>% 
    summarise(cnt = sum(in_place)) %>%
  ggplot(aes(x = reorder(description, cnt), y = cnt)) + 
    geom_bar(stat = 'identity', color='black', fill = '#d3d3d3') +
    coord_flip() +
    lims(y = c(0,8)) +
    ylab('# respondents') +
    xlab('task') +
    ggtitle(paste0('figure ', figure_n, ': tasks fulfilled by consultants for 8 managers at firms in Belgium')) +
    theme_tufte() +
    theme(
      plot.title = element_text(size=10),
      axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0), size = 10),
      axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0), size = 10),
      plot.title.position = 'plot'
      )
```

## Rationale (RQ2)

There is a clear pattern when it comes to the rationale for hiring consultants. They have the skills that are not available in-house, or they are hard to recruit. Furthermore, they're the experts in the tasks that they are hired for, and they can look at it with an outsider perspective.

Respondent 3 explicitly mentioned that a consultant is their soundboard: the respondent has the inside knowledge, and the consultants has knowledge about the task at hand. A back-and-forth with a consultants is an iteration to come to the best solution for the company. Respondent 1 agreed. Consultants are good at tasks with low asset specificity. Without the necessary commitment, it's impossible to grasp the company. Only employees, with long-term commitment, succeed at grasping the complex product of the company.

Multiple respondents mentioned how hard it is today to hire people for specific tasks, and for which they are forced to rely on consultants. They claim that this is a Belgian problem: tech-savvy people are seduced with a higher wage and fringe benefits that only consultancy firms can offer. Furthermore, lots of senior experts become freelance consultants due to its fiscally attractive regime.

The topic of cost control and/or cost reduction sparked discussion with three respondents. For respondent 1, hiring consultants vis--vis recruiting new staff meant an internalization of costs and that it was added to the P&L. Respondent 2 explained how consultants were often chosen as it was easy to assess beforehand what the cost would be. Finally, respondent 7 described how many managers preferred consultants because they don't come with indirect costs like unionization & pension scheme obligations.

$~$

```{r, echo=FALSE, out.width="75%", fig.align="center"}
figure_n <- figure_n + 1
data %>% 
  select(all_of('ID') | starts_with("Q6")) %>%
  pivot_longer(cols = starts_with("Q6"), names_to = 'reason', values_to = 'mentioned') %>%
  left_join(legend, by = join_by(reason == Question)) %>%
  mutate(description = paste(reason, Description, sep = ': ')) %>%
  select(all_of(c('ID', 'description', 'mentioned'))) %>%
  group_by(description) %>% 
    summarise(cnt = sum(mentioned)) %>%
  ggplot(aes(x = reorder(description, cnt), y = cnt)) + 
    geom_bar(stat = 'identity', color='black', fill = '#d3d3d3') +
    coord_flip() +
    lims(y = c(0,8)) +
    ylab('# respondents') +
    xlab('task') +
    ggtitle(paste0('figure ', figure_n, ': reasons for using consultants according to 8 managers at firms in Belgium')) +
    theme_tufte() +
    theme(
      plot.title = element_text(size=10),
      axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0), size = 10),
      axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0), size = 10),
      plot.title.position = 'plot'
      )
```

## Success (RQ3)

A clear picture emerged when asking the respondents about how they define and evaluate the success of a consultancy engagement. Most respondents have no formal, objective way for determining if a consultancy engagement was successful. Respondent 1 described how there are statements of work, but no statements of success. When problems arise during a consultancy assignment, this results in a mutual blame game. Respondent 5 admitted they never even considered what defines a successful consultancy engagement.

Two informal definitions of success did, however, arise during the follow-up discussion for this question

-   The job is done.

-   The respondent and the consultant hit it off.

The first is objective, yet almost tautological. Furthermore, it does not account for any kinds of moral hazard that could have occurred during the engagement. Finally, it does not take into account the opportunity costs when a task could have been done internally.

The second is completely subjective, and hard to rationalize. This hard-to-define bond between a manager and a consultants shows that there's at least some truth in the more critical theoretical frameworks, like social identity theory.

## Agency Problems (RQ4)

### Adverse selection

Only one respondent claimed to have never encountered adverse selection with a consultants. That means that seven respondents have ever hired a consultant who did not have the adequate skill set to complete one or more tasks. Respondent 1 described how the mismatch can be related to both hard skills or cultural fit.

Respondent 2 and 3 described how many firms overpromise on the skills of their consultants. Respondent 6 described how adverse selection is rampant with near-shore implementation consultants, but that the good consultants definitely make up for the bad consultants. Respondent 7 made an interesting remark: "I'm not sure they do it intentionally. Sometimes they seem to overestimate their own skills."

### Moral Hazard

Six respondents described how they have encountered types of moral hazard when working with consultants.

-   Lack of passion. Respondent 1 elaborated on consultants not taking the time to understand the context, and instead focusing to hard on the process. The result is that they don't get traction, don't get heard, become demotivated and quickly exit.
-   Training a consultant. Respondent 4 described how the organization often pays the dues for training a junior consultant, and it takes quite a lot of time to realize that this is happening. This is closely related to adverse selection.
-   Slacking off: Respondent 6 described how consultants have slacked off before. An important enabler for this behavior is working remotely.

No respondent described a conflict of interest or unethical behavior.

$~$

```{r, echo=FALSE, out.width="75%", fig.align="center"}
figure_n <- figure_n + 1
data %>% 
  select(all_of('ID') | starts_with("Q7") | starts_with("Q10")) %>%
  pivot_longer(cols = starts_with("Q7") | starts_with("Q10"), names_to = 'answer', values_to = 'count') %>%
  left_join(legend, by = join_by(answer == Question)) %>%
  mutate(description = paste(answer, Description, sep = ': ')) %>%
  select(all_of(c('ID', 'description', 'count'))) %>%
  group_by(description) %>% 
    summarise(cnt = sum(count)) %>%
  ggplot(aes(x = reorder(description, -cnt), y = cnt)) + 
    geom_bar(stat = 'identity', color='black', fill = '#d3d3d3') +
    coord_flip() +
    lims(y = c(0,8)) +
    ylab('# respondents') +
    xlab('task') +
    ggtitle(paste0('figure ', figure_n, ': experience with agency problems according to 8 managers at firms in Belgium')) +
    theme_tufte() +
    theme(
      plot.title = element_text(size=10),
      axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0), size = 10),
      axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0), size = 10),
      plot.title.position = 'plot'
      )

```

## Controlling Adverse Selection (RQ5)

From the respondents' answers, we can confidently state that most organizations have at least some purchase regulation, or involvement from the procurement department. Most repondents also indicated that reputation played some kind of role for the selection of consultancy firms and even individual consultants.

### Reputation

Respondent 1 described how there are formal reputational checks in place for consultants, and that this is outsourced to the partnering organization that recruits freelance consultants. When working directly with consultancy firms, reputational checks are done in-house.

According to respondent 2, the procurement department at their organization values reputation a lot. Any consultancy firm that wants to bring in consultants us subjected to reputation checks. For individual consultants, this is rather done informally. Respondents 4, 6 and 7agree, and describe that internal reputation is very important for consultancy firms to end up on the short list to which the RFP is sent.

### Procurement

Respondent 1 described that there is no obligation to involve the procurement department. However, for contracts over 250.000 euro, they can voluntarily order contractual reviews & advice with the procurement department.

The organization of respondent 2 makes a distinction between consultants that are hired for project, and those for roles for which they can't find the required profile. Procurement is mostly involved for project-based consultants, while HR is responsible for role-based consultants. The organization also works with a managed service provider (MSP), for which procurement sets the expected fees and preferred suppliers. Based on input from the manager, the MSP will look for consultants (both freelance and from consultancy firms). Individual consultants have to do an intake with both HR and the hiring manager.

Respondents 5 and 7 explained that procurement is involved for the legal and financial side of a consultancy engagement. They also have a list of preferred suppliers.

For large consultancy firms, the organization of respondent 6 has a preferred supplier list. Consultants with a niche expertise and freelance consultants need to go through a managed service provider.

### Purchase regulation

Respondent 1, 7 and 8 explained the various price forks for consultancy contracts. Depending on the order of magnitude of the contract, other procedures applied. While purchase regulation is mostly related to the chain of approval, and aren't very likely to prevent adverse selection, there is also some in-house expertise with regards to acceptable fees and compensation with respondent 1.

The consultancy budgets in the organization of respondent 2 are assigned to each director. When a mid-level manager needs consultants, the director has to sign off. This is more about controlling the manager and is unlikely to catch adverse selection.

Respondent 6 described how consultancy firms need the required insurance clauses and should have done the necessary business process risk assessments. While this could be a proxy for the quality of a consultancy firm, this is unlikely to catch poor individual consultants.

### Third-party assessments

Respondent 1 described how consultants that are expected to lead people are subjected to a third-party assessment that polls for cultural fit and managerial skills.

### Consulting database

The organization of respondent 3 does not have controls in place to prevent adverse selection. However, they describe having a personal list of preferred consultants and firms. Other managers also have such a list. Respondent 8 agreed.

```{r, echo=FALSE, out.width="75%", fig.align="center"}
figure_n <- figure_n + 1
data %>% 
  select(all_of('ID') | starts_with("Q8")) %>%
  pivot_longer(cols = starts_with("Q8"), names_to = 'control_mech', values_to = 'in_place') %>%
  filter(in_place != 0) %>%
  mutate(in_place = as.character(in_place)) %>%
  mutate(in_place = str_replace(in_place, '1', 'informal')) %>%
  mutate(in_place = str_replace(in_place, '2', 'formal')) %>%
  mutate(in_place = fct_relevel(in_place, 'informal', 'formal')) %>%
  left_join(legend, by = join_by(control_mech == Question)) %>%
  mutate(description = paste(control_mech, Description, sep = ': ')) %>%
  select(all_of(c('ID', 'description', 'in_place'))) %>%
  group_by(description, in_place) %>% 
    summarise(cnt = n()) %>%
  ggplot(aes(x = reorder(description, cnt), y = cnt, fill = in_place)) + 
    geom_bar(stat = 'identity', color='black') +
    scale_fill_manual(values = c('#d3d3d3', '#000000'), name = '') +
    coord_flip() +
    lims(y = c(0,8)) +
    ylab('# respondents') +
    xlab('task') +
    ggtitle(paste0('figure ', figure_n, ': mechanisms for controlling adverse selection with consultants according to 8 managers at firms in Belgium')) +
    theme_tufte() +
    theme(
      plot.title = element_text(size=10),
      axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0), size = 10),
      axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0), size = 10),
      plot.title.position = 'plot',
      legend.position = 'bottom'
      )

```

## Controlling Moral Hazard (RQ6)

### First-party assessments

Respondent 1 described that for most consultants, there is a standard selection procedure, but excluding the HR-component.

For respondent 2, interviews are extremely important and are used to assess if a consultant will fit in at the company. They also stated not to be surprised if managers have hired consultants without having any knowledge if that consultant has the capabilities to execute the task, or will fit in with the team. Respondent 6 agrees. Due to the many mismatched with on-shore and off-shore consultants, interviews are crucial.

### Monitoring

According to respondent 1, consultants are monitored in the same way as consultants. There are recurrent 1-on-1s and recurring status updates However, a consultant's goals are defined more clearly, with the given task in mind. For this reason, they are evaluated much more stringently. Nevertheless, the respondent concurred that a lot depends on the hiring manager. Respondent 3, 4 and 5 agree and have recurrent status updates with their consultants.

### Whistleblowing

Respondent 2 indicated that they expect employees to reports consultants that underperform, and occasionally polls for opinions.

### Incentivization

Respondent 1 argued that the continuation of the contract should be an incentive an sich, although formal incentivization has been used occasionally in the past. At the company of respondent 6, contractual clauses exist that reward bonuses or refunds. However, the respondent doubted that this has any impact on the individual consultants. Respondent 8 confirms that consultancy firms can charge 90% of their price and an additional 10% if they meet deadlines or targets. The organization of respondent 4 allows consultancy firms to make a case for their website, in exchange for better contract terms.

### Third-party moderation

At the company of respondent 1, the management of projects and the actual execution of a project are handed to two different parties for large key projects. Respondent 7 has worked with consultants from multiple firms on a single project and assumes that they would call each other out when moral hazard occurs. Respondents 4, 6 and 8 described that a third-party has been involved to perform code audits, or to deescalate a projects when a rift between the organization and the consultancy firm has emerged.

### Relational

Within the team of respondent 1, consultants are treated in the same way as payroll workers. Nevertheless, a lot depends on the hiring manager. Respondent 5 agrees.

For respondent 2 this is an active (yet informal) policy: they expect consultants to integrate neatly with the team, and are invited to the same team building activities as staff. They believe that this is both in the best interest of the company and of the individual consultant. This is very different from the formal company policy, that wants to create a distance between the consultants and the company, for example by giving their badge lanyards a different color. Consultants are also not invited to company events, unless they pay for their entrance ticket (which is 100 euro). Respondent 6 concurs: their senior management are also wary of consultants and don't expect managers to invest time in them.

At the company of respondent 3 the same rules apply: consultants are identified by a different lanyard, and are not invited to company meetings. They are however welcome at team meetings. Respondent 4 agrees: consultants can attend team meetings, but it would be extremely frowned upon if consultants bill that time.

$~$

```{r, echo=FALSE, out.width="75%", fig.align="center"}
figure_n <- figure_n + 1
suppressMessages(data %>% 
  select(all_of('ID') | starts_with("Q11")) %>%
  pivot_longer(cols = starts_with("Q11"), names_to = 'control_mech', values_to = 'in_place') %>%
  filter(in_place != 0) %>%
  mutate(in_place = as.character(in_place)) %>%
  mutate(in_place = str_replace(in_place, '1', 'informal')) %>%
  mutate(in_place = str_replace(in_place, '2', 'formal')) %>%
  mutate(in_place = fct_relevel(in_place, 'informal', 'formal')) %>%
  left_join(legend, by = join_by(control_mech == Question)) %>%
  mutate(description = as.factor(paste(control_mech, Description, sep = ': '))) %>%
  mutate(description = fct_rev(fct_relevel(description, 
                                   'Q11f: relational', 
                                   'Q11a: first-party assessments', 
                                   'Q11e: third-party moderation', 
                                   'Q11b: monitoring',
                                   'Q11d: incentivization', 
                                   'Q11c: whistleblowing'
                                   ))) %>%
  select(all_of(c('ID', 'description', 'in_place'))) %>%
  group_by(description, in_place) %>% 
    summarise(cnt = n()) %>%
  ggplot(aes(x = description, y = cnt)) + 
    geom_bar(stat = 'identity', color='black', aes(fill = in_place)) +
    scale_fill_manual(values = c('#d3d3d3', '#000000'), name = '') +
    coord_flip() +
    lims(y = c(0,8)) +
    ylab('# respondents') +
    xlab('task') +
    ggtitle(paste0('figure ', figure_n, ': mechanisms for controlling moral hazard with consultants according to 8 managers at firms in Belgium')) +
    theme_tufte() +
    theme(
      plot.title = element_text(size=10),
      axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0), size = 10),
      axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0), size = 10),
      plot.title.position = 'plot'
      ))

```

# Limitations

# Conclusions

# References
